---
title: "Paxton Boyd"
---

## Data and Setup

### Packages

```{r}
#| label: setup

library(tidyverse)
library(scales)
library(ggridges)
library(patchwork)
library(boot)          # bootstrap Gini confidence intervals
library(quantreg)      # quantile regression
library(broom)         # tidy model outputs
library(gt)            # publication-quality tables
library(modelsummary)  # regression summary tables
library(ggtext)        # markdown / HTML in ggplot labels
library(glue)          # string interpolation

# ── Shared palette ─────────────────────────────────────────────────────────────
pal_size <- c("Micro" = "#e76f51", "Small" = "#f4a261", "Medium/Large" = "#264653")
pal_gov  <- c(
  "Sole Entrepreneur"                    = "#e9c46a",
  "Managing Partner /\nAdministrator"    = "#4a90d9",
  "Beneficial Owner\n(passive/investor)" = "#264653"
)

caption_base <- "Source: Brazilian CNPJ Registry ·He, Miao & Boyd"
```


### Data Ingestion and Cleaning

```{r}
#| label: data-load

companies <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv",
  show_col_types = FALSE
)

# ── Core recoding ───────────────────────────────────────────────────────────────
co <- companies |>
  # Drop records with missing or non-positive capital (not usable for log-scale analysis)
  filter(
    !is.na(company_size),
    !is.na(capital_stock),
    capital_stock > 0,
    !is.na(owner_qualification)
  ) |>
  mutate(
    # Size tier (ordered factor)
    size = case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE                               ~ "Medium/Large"
    ),
    size = factor(size, levels = c("Micro", "Small", "Medium/Large")),

    # Governance archetype: collapse owner_qualification into 4 classes
    governance = case_when(
      str_detect(owner_qualification,
                 regex("beneficial owner", ignore_case = TRUE)) ~
        "Beneficial Owner\n(passive/investor)",
      str_detect(owner_qualification,
                 regex("entrepreneur|business owner", ignore_case = TRUE)) ~
        "Sole Entrepreneur",
      str_detect(owner_qualification,
                 regex("managing partner|administrator", ignore_case = TRUE)) ~
        "Managing Partner /\nAdministrator",
      TRUE ~ "Other"
    ),
    governance = factor(governance, levels = c(
      "Sole Entrepreneur",
      "Managing Partner /\nAdministrator",
      "Beneficial Owner\n(passive/investor)",
      "Other"
    )),

    # Log10 capital for distributional analysis
    log_capital = log10(capital_stock),

    # Legal-form shorthand for micro analysis
    legal_short = case_when(
      str_detect(legal_nature, "LLC")                 ~ "LLC",
      str_detect(legal_nature, "Sole Proprietorship") ~ "Sole Prop.",
      str_detect(legal_nature, "Individual Limited")  ~ "EIRELI",
      str_detect(legal_nature, "Simple")              ~ "Simple Part.",
      str_detect(legal_nature, "Silent")              ~ "Silent Part.",
      str_detect(legal_nature, "Corporation")         ~ "Corporation",
      TRUE                                            ~ "Other"
    ),

    # Capital-rich flag: top quartile *within* micro-enterprises
    capital_q75_micro     = quantile(capital_stock[size == "Micro"], 0.75, na.rm = TRUE),
    is_capital_rich_micro = (size == "Micro") & (capital_stock >= capital_q75_micro)
  )

cat("Final observations:", scales::comma(nrow(co)), "\n")
```

### Descriptive Summary Table

```{r}
#| label: summary-table

co |>
  filter(governance != "Other") |>
  group_by(size, governance) |>
  summarise(
    n        = n(),
    med_cap  = median(capital_stock),
    mean_cap = mean(capital_stock),
    p25_cap  = quantile(capital_stock, 0.25),
    p75_cap  = quantile(capital_stock, 0.75),
    .groups  = "drop"
  ) |>
  gt(groupname_col = "size") |>
  tab_header(
    title    = md("**Capital Stock by Firm Size and Governance Archetype**"),
    subtitle = "All values in BRL · Median, mean, and interquartile range"
  ) |>
  cols_label(
    governance = "Governance Archetype",
    n          = "N",
    med_cap    = "Median",
    mean_cap   = "Mean",
    p25_cap    = "P25",
    p75_cap    = "P75"
  ) |>
  fmt_number(
    columns  = c(med_cap, mean_cap, p25_cap, p75_cap),
    suffixing = TRUE, decimals = 1
  ) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) |>
  tab_options(
    table.font.size            = px(13),
    heading.align              = "left",
    row_group.background.color = "#f4f4f4"
  ) |>
  tab_source_note(md("*Source: Brazilian CNPJ Registry*"))
```

---
## Analysis I — Within-Tier Inequality {#sec-inequality}

### Lorenz Curves with Bootstrapped Gini Coefficients

A Gini coefficient point estimate is not sufficient for inference: it has no standard error as
typically presented. We bootstrap 999 resamples per size tier to construct 95% percentile
confidence intervals, confirming whether the ordering across tiers is statistically reliable.

```{r}
#| label: gini-bootstrap
#| cache: true

# ── Gini helper ─────────────────────────────────────────────────────────────────
gini <- function(x) {
  x <- sort(x[x > 0 & !is.na(x)])
  n <- length(x)
  if (n < 2L) return(NA_real_)
  2 * sum(seq_len(n) * x) / (n * sum(x)) - (n + 1L) / n
}

gini_stat <- function(data, i) gini(data[i])

set.seed(112)
gini_ci <- co |>
  group_by(size) |>
  group_modify(~ {
    b  <- boot(.x$capital_stock, gini_stat, R = 999)
    ci <- boot.ci(b, type = "perc", conf = 0.95)
    tibble(
      g     = b$t0,
      lower = ci$percent[4L],
      upper = ci$percent[5L],
      n     = nrow(.x)
    )
  }) |>
  ungroup()

gini_ci |>
  mutate(across(c(g, lower, upper), ~ round(.x, 4))) |>
  gt() |>
  tab_header(
    title    = md("**Bootstrapped Gini Coefficients by Firm Size**"),
    subtitle = "999 resamples · 95% percentile confidence interval"
  ) |>
  cols_label(
    size  = "Tier", g = "Gini",
    lower = "CI Lower", upper = "CI Upper", n = "N"
  ) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0)
```

```{r}
#| label: lorenz-plot
#| fig-width: 10
#| fig-height: 5.5

lorenz_data <- co |>
  group_by(size) |>
  arrange(capital_stock, .by_group = TRUE) |>
  mutate(
    cum_firms   = row_number() / n(),
    cum_capital = cumsum(capital_stock) / sum(capital_stock)
  ) |>
  ungroup()

gini_labels <- gini_ci |>
  mutate(
    label = glue("Gini = {round(g, 3)}\n[{round(lower, 3)}, {round(upper, 3)}]"),
    x = 0.20,
    y = c(0.70, 0.59, 0.48)
  )

ggplot(lorenz_data, aes(x = cum_firms, y = cum_capital, colour = size)) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", colour = "grey60", linewidth = 0.55) +
  geom_line(linewidth = 1.1, alpha = 0.9) +
  geom_ribbon(
    data    = filter(lorenz_data, size == "Micro"),
    aes(ymin = cum_capital, ymax = cum_firms),
    fill    = "#e76f51", alpha = 0.07, colour = NA
  ) +
  geom_label(
    data        = gini_labels,
    aes(x = x, y = y, label = label, colour = size),
    inherit.aes = FALSE,
    fill        = "white", label.size = 0.25,
    fontface    = "bold", size = 3.3, lineheight = 1.2
  ) +
  annotate("text", x = 0.59, y = 0.48,
           label = "Perfect equality", colour = "grey55",
           size = 3.2, angle = 42) +
  scale_colour_manual(values = pal_size, name = "Firm size") +
  scale_x_continuous(labels = percent_format(),
                     name   = "Cumulative share of firms") +
  scale_y_continuous(labels = percent_format(),
                     name   = "Cumulative share of capital") +
  labs(
    title   = "Micro-enterprises are internally the most unequal — **Gini > 0.60**",
    subtitle = glue(
      "Lorenz curves by firm size tier. Labels show Gini point estimate and ",
      "bootstrapped 95% CI (999 resamples).\n",
      "The further the curve bows below the diagonal, the more concentrated ",
      "capital is among a small fraction of firms within that tier."
    ),
    caption = caption_base
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_markdown(face = "bold", size = 14, colour = "#264653"),
    plot.subtitle    = element_text(colour = "grey40", size = 10.5,
                                    lineheight = 1.3, margin = margin(b = 14)),
    plot.caption     = element_text(colour = "grey50", size = 8, hjust = 0),
    legend.position  = "top",
    panel.grid.minor = element_blank()
  )
```

The bootstrapped confidence intervals confirm that the Gini point estimates are precise: the
non-overlapping 95% CIs establish that all three tiers differ significantly in their degree of
internal inequality. Micro-enterprises (Gini > 0.60) are substantially more internally
unequal than either small or medium/large firms.

### Theil T Decomposition: Within vs. Between Groups

The Lorenz curve compares tiers but cannot partition total inequality into *within-tier* and
*between-tier* components. The **Theil T index** (GE(1)) supports an exact additive
decomposition:

$$T = T_{\text{between}} + T_{\text{within}} = \sum_g \frac{n_g}{N}\frac{\mu_g}{\mu}\ln\frac{\mu_g}{\mu} \;+\; \sum_g \frac{n_g}{N}\frac{\mu_g}{\mu} T_g$$
where $T_g$ is the within-group Theil T for group $g$, $n_g$ is the group count, $\mu_g$ is the
group mean capital, and $\mu$ is the overall mean.

```{r}
#| label: theil-decomp
#| fig-width: 9
#| fig-height: 4.5

theil_t <- function(x) {
  x  <- x[x > 0 & !is.na(x)]
  mu <- mean(x)
  mean((x / mu) * log(x / mu))
}

decomp <- co |>
  group_by(size) |>
  summarise(
    n_g = n(),
    mu_g = mean(capital_stock),
    T_g  = theil_t(capital_stock),
    .groups = "drop"
  ) |>
  mutate(
    N             = sum(n_g),
    mu            = sum(n_g * mu_g) / N,
    w_g           = (n_g / N) * (mu_g / mu),
    T_between_g   = w_g * log(mu_g / mu),
    T_within_g    = w_g * T_g
  )

T_between <- sum(decomp$T_between_g)
T_within  <- sum(decomp$T_within_g)
T_total   <- T_between + T_within

# Panel A: total within vs between
summary_bar <- tibble(
  component = c("Between tiers", "Within tiers"),
  value     = c(T_between, T_within),
  pct       = c(T_between / T_total, T_within / T_total)
)

p_summary <- ggplot(summary_bar, aes(x = component, y = pct, fill = component)) +
  geom_col(width = 0.55, alpha = 0.9) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)),
            vjust = -0.5, fontface = "bold", size = 4.2) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1),
                     name = "Share of total Theil T") +
  scale_fill_manual(
    values = c("Between tiers" = "#264653", "Within tiers" = "#e76f51"),
    guide  = "none"
  ) +
  labs(title    = "Total decomposition",
       subtitle = glue("Total Theil T = {round(T_total, 3)}"),
       x = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", colour = "#264653"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank())

# Panel B: within-tier contributions broken down by size
p_within <- decomp |>
  mutate(pct_of_total = T_within_g / T_total) |>
  ggplot(aes(x = fct_rev(size), y = pct_of_total, fill = size)) +
  geom_col(width = 0.55, alpha = 0.9) +
  geom_text(aes(label = percent(pct_of_total, accuracy = 0.1)),
            vjust = -0.5, fontface = "bold", size = 4.2) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.7),
                     name = "Within-tier share of total Theil T") +
  scale_fill_manual(values = pal_size, guide = "none") +
  labs(
    title    = "Within-tier breakdown by size",
    subtitle = "Each bar = one tier's contribution to the within-group component",
    x = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", colour = "#264653"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank())

p_summary + p_within +
  plot_annotation(
    title   = "Most of Brazil's capital inequality is *within* size tiers, not between them",
    subtitle = glue(
      "Theil T decomposition. Within-tier component = {percent(T_within / T_total, 0.1)} ",
      "of total inequality — the largest driver is inequality inside micro-enterprises."
    ),
    caption = caption_base,
    theme   = theme(
      plot.title    = element_markdown(face = "bold", size = 14, colour = "#264653"),
      plot.subtitle = element_text(colour = "grey40", size = 10.5, lineheight = 1.3),
      plot.caption  = element_text(colour = "grey50", size = 8)
    )
  )
```

The decomposition is the key structural result of this section. If most inequality were
*between* tiers, then adjusting the revenue classification cutoffs could reduce aggregate
inequality. But if the dominant component is *within* tiers, revenue-threshold reform achieves
little: the same cutoff that separates micro from small cannot distinguish capital-poor
subsistence firms from well-endowed holding-company subsidiaries within the micro category.
The within-tier dominance directly motivates the governance analysis that follows.

---
