---
title: "Capital Without Category — Part II"
subtitle: "Governance as distributional shape: owner archetype and the heavy-tail signature"
author: "Jerry Miao"
date: "2026-02-25"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-dpi: 180
    fig-align: center
    fig-cap-location: bottom
    df-print: paged
execute:
  warning: false
  message: false
  cache: true
---

## Background

Part I established that capital concentration in Brazilian formal-sector firms is a
universal property: every firm-size tier exhibits Gini > 0.98 and a top-1% capital
share of 96–99%. The shape of the distribution is the same across regulatory
categories. What differs across tiers is *who occupies the top* — and the firm-size
classification cannot answer that question, because it is built on annual revenue and
is silent on the legal architecture of the controlling party.

The CNPJ registry's `owner_qualification` field carries that missing information. It
records the legal role of the controlling person under three broad archetypes:

- **Sole Entrepreneur** — a natural person operating under their own name, fully
  exposed to business risk.
- **Managing Partner / Administrator** — an active operator who runs the business as
  one of one or more named partners with operational authority.
- **Beneficial Owner (passive / investor)** — a controlling equity holder who may sit
  behind a holding-company chain, often without operational involvement.

These archetypes map onto well-established predictions from agency theory. Sole
Entrepreneurs face the highest information asymmetries with external lenders and have
no separation between personal and business liability. Managing Partners signal active
human capital and shared risk-bearing, lowering the information cost of credit.
Beneficial Owners are heterogeneous: at the small end they are dormant shells; at the
large end they are investment vehicles channelling capital from institutional or
family-office structures.

## Research Question

> **Q2.** Conditional on firm size, does governance archetype shape the capital
> distribution? If it does, is the effect a *level shift* (governance moves the
> median), a *variance shift* (governance moves the tail), or both?

The question matters because the answer dictates the targeting strategy. A level
shift means governance acts as a coarse credit-access proxy. A variance shift means
governance is selecting different *populations* of firms — and lumping them by size
will systematically misallocate any size-targeted policy.

## Design Conventions

```{r}
#| label: setup-theme

library(tidyverse)
library(scales)
library(ggridges)
library(gt)

# ── Shared palette ──────────────────────────────────────────────────────────
pal_size <- c(
  "Micro"        = "#c3553a",
  "Small"        = "#d9a441",
  "Medium/Large" = "#2a5a6b"
)

pal_gov <- c(
  "Sole Entrepreneur"                    = "#d9a441",   # amber (solo)
  "Managing Partner /\nAdministrator"    = "#5b8a9a",   # slate blue (active)
  "Beneficial Owner\n(passive/investor)" = "#1d3557"    # deep navy (passive)
)

# ── Shared theme ────────────────────────────────────────────────────────────
theme_cwc <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title        = element_text(face = "bold", size = base_size + 2,
                                       colour = "#1d3557", margin = margin(b = 4)),
      plot.subtitle     = element_text(colour = "grey35", size = base_size - 2,
                                       lineheight = 1.35, margin = margin(b = 18)),
      plot.caption      = element_text(colour = "grey55", size = base_size - 5,
                                       hjust = 0, margin = margin(t = 12)),
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      strip.text        = element_text(face = "bold", size = base_size - 1,
                                       colour = "#1d3557"),
      strip.background  = element_rect(fill = "grey96", colour = NA),
      panel.grid.minor  = element_blank(),
      panel.grid.major  = element_line(colour = "grey92", linewidth = 0.3),
      panel.spacing     = unit(1.4, "lines"),
      legend.position   = "top",
      legend.title      = element_text(face = "bold", size = base_size - 2),
      legend.text       = element_text(size = base_size - 2),
      axis.title        = element_text(colour = "grey30", size = base_size - 2),
      axis.text         = element_text(colour = "grey40", size = base_size - 3)
    )
}

caption_text <- paste(
  "Source: Brazilian CNPJ Registry · TidyTuesday 2026 Wk 4 (Extended)",
  "· Boyd, He & Miao · 2026-02-25"
)
```

## Data Preparation

```{r}
#| label: data-prep

companies <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv",
  show_col_types = FALSE
)

co <- companies |>
  filter(
    !is.na(company_size),
    !is.na(capital_stock),
    !is.na(owner_qualification),
    capital_stock > 0
  ) |>
  mutate(
    size = case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE                               ~ "Medium/Large"
    ),
    size = factor(size, levels = c("Micro", "Small", "Medium/Large")),

    governance = case_when(
      str_detect(owner_qualification,
                 regex("beneficial owner", ignore_case = TRUE))     ~ "Beneficial Owner\n(passive/investor)",
      str_detect(owner_qualification,
                 regex("entrepreneur|business owner", ignore_case = TRUE)) ~ "Sole Entrepreneur",
      str_detect(owner_qualification,
                 regex("managing partner|administrator", ignore_case = TRUE)) ~ "Managing Partner /\nAdministrator",
      TRUE ~ "Other"
    ),
    governance = factor(governance, levels = c(
      "Sole Entrepreneur",
      "Managing Partner /\nAdministrator",
      "Beneficial Owner\n(passive/investor)",
      "Other"
    )),

    log_capital = log10(capital_stock)
  )

cat("Observations:", comma(nrow(co)), "\n")
cat("Governance archetypes:\n"); print(count(co, governance, sort = TRUE))
```

## Descriptive Snapshot: Governance × Size Cells

The table below carries the data we will read off the ridges. The mean-to-median ratio
is the single statistic worth watching: when this ratio is close to 1, the
distribution is roughly symmetric; when it is large (here, in the double digits for
Beneficial Owners), the mean is being pulled by a heavy right tail.

```{r}
#| label: tbl-governance-cells

gov_cells <- co |>
  filter(governance != "Other") |>
  group_by(size, governance) |>
  summarise(
    n         = n(),
    median    = median(capital_stock),
    mean      = mean(capital_stock),
    p90       = quantile(capital_stock, 0.90),
    mean_med  = mean / median,
    .groups   = "drop"
  )

gov_cells |>
  mutate(governance = str_replace(governance, "\n", " ")) |>
  gt(groupname_col = "size") |>
  tab_header(
    title    = md("**Capital stock by size × governance**"),
    subtitle = "Mean / median ratio diagnoses tail heaviness within each cell"
  ) |>
  cols_label(
    governance = "Governance archetype",
    n = "N", median = "Median", mean = "Mean",
    p90 = "P90", mean_med = "Mean / Median"
  ) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = c(median, mean, p90),
             suffixing = TRUE, decimals = 1) |>
  fmt_number(columns = mean_med, decimals = 1) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) |>
  tab_options(
    table.font.size            = px(13),
    heading.align              = "left",
    row_group.background.color = "#f4f4f4"
  ) |>
  tab_source_note(md("*Source: Brazilian CNPJ Registry*"))
```

## Analysis: Governance Shapes Distributions, Not Just Levels

### Methodological motivation

A boxplot or a bar chart of medians would already show that governance archetypes
differ on average — but it would miss the analytically critical question: do they
differ in *level* (a translation of the entire distribution) or in *shape* (different
variances and tail heaviness)? Density ridges with explicit median lines and mean
markers are the right diagnostic, because they let the reader see all three moments
on one plot:

- The **median** (white vertical line) is a robust level summary — pulled around
  much less by extreme observations than the mean.
- The **mean** (red tick) is sensitive to the tail.
- The **gap between median and mean**, read directly off the plot, signals tail
  heaviness without computing a moment.

For inferential support we use the **Kruskal-Wallis test** within each size tier.
This is a non-parametric one-way analysis of variance on ranks: appropriate here
because capital-stock distributions are extremely log-normal at best and have heavy
right tails that violate ANOVA's normality assumption. Kruskal-Wallis tests the null
hypothesis that all governance subgroups within a size tier are drawn from the same
distribution. Rejecting the null does not by itself say *what* differs — that
interpretation comes from the visual.

### Kruskal-Wallis tests within each size tier

```{r}
#| label: kw-test

kw_results <- co |>
  filter(governance != "Other") |>
  group_by(size) |>
  summarise(
    statistic = kruskal.test(capital_stock ~ governance)$statistic,
    df        = kruskal.test(capital_stock ~ governance)$parameter,
    p_value   = kruskal.test(capital_stock ~ governance)$p.value,
    n         = n(),
    .groups   = "drop"
  )

kw_results |>
  gt() |>
  tab_header(
    title    = md("**Kruskal-Wallis tests of governance × capital within each tier**"),
    subtitle = "H₀: all three governance subgroups draw from the same capital distribution"
  ) |>
  cols_label(
    size = "Tier", statistic = "H statistic",
    df = "df", p_value = "p-value", n = "N"
  ) |>
  fmt_number(columns = c(statistic), decimals = 1) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0) |>
  fmt_scientific(columns = p_value, decimals = 2) |>
  tab_options(table.font.size = px(13), heading.align = "left")
```

The null is rejected at conventional levels in every tier. Governance and capital
are not independent within any size category. The visual now supplies the *what*.

### Density ridges with median and mean markers

```{r}
#| label: fig-governance-ridges
#| fig-width: 12
#| fig-height: 6.2
#| fig-cap: "Governance shapes the capital distribution differently across archetypes."

gov_stats <- co |>
  filter(governance != "Other") |>
  group_by(size, governance) |>
  summarise(
    median_cap = median(capital_stock),
    mean_cap   = mean(capital_stock),
    p90_cap    = quantile(capital_stock, 0.90),
    n          = n(),
    .groups    = "drop"
  ) |>
  mutate(
    med_lab = paste0("median ", number(median_cap, scale = 1e-3, suffix = "k",
                                       big.mark = ",", accuracy = 1))
  )

ggplot(
  co |> filter(governance != "Other"),
  aes(x = log_capital, y = governance, fill = governance)
) +
  geom_density_ridges(
    alpha = 0.85, scale = 0.92, rel_min_height = 0.01,
    quantile_lines = TRUE, quantiles = 2,
    colour = "white", linewidth = 0.45
  ) +
  # mean marker (red tick) — surfaces the heavy-tail of Beneficial Owners
  geom_segment(
    data = gov_stats,
    aes(x = log10(mean_cap), xend = log10(mean_cap),
        y = as.numeric(governance) - 0.02,
        yend = as.numeric(governance) + 0.38),
    inherit.aes = FALSE,
    colour = "#c3553a", linewidth = 0.8
  ) +
  # median label placed above the ridge
  geom_text(
    data = gov_stats,
    aes(x = log10(median_cap), y = as.numeric(governance) + 0.72,
        label = med_lab),
    inherit.aes = FALSE,
    size = 3.0, colour = "grey25", hjust = 0.5
  ) +
  facet_wrap(~size, nrow = 1) +
  scale_fill_manual(values = pal_gov, guide = "none") +
  scale_x_continuous(
    breaks = c(4, 5, 6, 7, 8),
    labels = c("10k", "100k", "1M", "10M", "100M"),
    name   = "Capital stock (BRL, log scale)",
    expand = expansion(mult = c(0.02, 0.04))
  ) +
  scale_y_discrete(expand = expansion(add = c(0.3, 1.1))) +
  labs(
    title    = "Managing Partners lead the median; Beneficial Owners carry the tail",
    subtitle = paste0(
      "Density ridges show capital-stock distribution by owner archetype within each size tier.\n",
      "White line = median  •  Red tick = mean  •  X-axis: log₁₀ scale"
    ),
    y        = NULL,
    caption  = caption_text
  ) +
  theme_cwc() +
  theme(panel.grid.major.y = element_blank())
```

### Interpretation

Reading the ridges row by row tells three different stories.

**Managing Partners — the level archetype.** Median capital scales with firm size:
roughly 300k at Micro, 360k at Small, **1,000k at Medium/Large**. This is the only
archetype whose median moves materially across tiers. The interpretation is that
active operators *accumulate* capital as firms grow — the median tracks size because
capital and operational scale grow together. The mean tick sits close to the median,
indicating a relatively well-behaved log-normal shape.

**Sole Entrepreneurs — the flat archetype.** Median capital is approximately 300k
across all three tiers; the entire distribution barely shifts. Sole Entrepreneurs
who are nominally classified as "Medium/Large" by revenue still carry roughly the
same capital as Sole Entrepreneurs classified as "Micro". The interpretation is
consistent with agency theory: solo proprietors face information asymmetries that
cap external capital access regardless of revenue scale. They are the only archetype
for which the size category does *not* signal capital endowment.

**Beneficial Owners — the heavy-tail archetype.** The median is low and flat
(~300k) across all three tiers, indistinguishable from Sole Entrepreneurs at
median. But the mean tick (red) sits far to the right of the median, and the
distribution shows a visible long right tail. The mean / median ratio in the
descriptive table reaches double digits — the statistical fingerprint of a
heavy-tailed mixture. Most Beneficial Owner-controlled firms are small (consistent
with shells, dormant equity holdings, family registries); a minority are very large
(consistent with holding-company subsidiaries and investment vehicles). The
distribution is bimodal in *type* even when it looks unimodal in *density*, because
the same regulatory category covers two separate populations.

The governance premium is not uniform. It is a **level premium** for Managing
Partners — they translate operational growth into capital — and a **variance
premium** for Beneficial Owners — they sort the population into a small upper tail
that the median misses. Sole Entrepreneurs receive neither. Any size-targeted
policy that treats these three archetypes as equivalent within a tier will
systematically misallocate.

This finding is the empirical foundation for the next question. If Beneficial
Owners carry the heavy tail, *which legal-form channel does this happen through*?
That is the subject of Part III.

## Methodological Notes

- **Why Kruskal-Wallis rather than ANOVA.** Capital-stock distributions are
  strongly right-skewed even after log transformation; ANOVA's normality assumption
  is violated. Kruskal-Wallis operates on ranks and is robust to this.
- **Why both median and mean on the same plot.** The two moments are most informative
  in *combination*: their gap is the heavy-tail signal. Reporting only the median
  would erase the Beneficial Owner story entirely; reporting only the mean would be
  pulled around by single firms.
- **Excluding "Other" governance.** A residual category with heterogeneous coding
  (state-owned representatives, foreign legal entities, etc.) is excluded from the
  ridges to keep archetype-level interpretation clean. It is approximately 6% of
  observations.
- **Why we dropped quantile regression here.** An earlier draft estimated
  τ ∈ {0.10, 0.25, 0.50, 0.75, 0.90} regressions. The ridge plot communicates the
  same finding (premium widens at the upper tail) more directly and without the
  rank-conditioned interpretation that quantile regression imposes. The
  distributional shape *is* the result.

## Forward Link

If Beneficial Owners carry the heavy tail of capital within every size tier, the
specific question for development policy is *which legal-form channel* turns that
ownership pattern into a registered micro-enterprise with above-tier-typical
capital. Part III ranks the legal-form × governance cells inside the micro tier and
identifies the partnership-with-passive-equity structure as the dominant escape
route — a result that overturns the popular "EIRELI-as-escape-route" framing.
