---
title: "Capital Without Category — Part I"
subtitle: "The universal power law: capital concentration across Brazilian firm-size tiers"
author: "Zhijun He"
date: "2026-02-25"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-dpi: 180
    fig-align: center
    fig-cap-location: bottom
    df-print: paged
execute:
  warning: false
  message: false
  cache: true
---

## Background

Brazil's official firm-size classification (Simples Nacional) is built on a single
threshold: **annual gross revenue**. A sole proprietor earning 360,000 BRL is legally a
"micro-enterprise" regardless of whether their declared capital stock is 50,000 BRL or
5,000,000 BRL. This regulatory choice creates a natural experiment in the CNPJ registry:
firms within the same legal category can have wildly different capital endowments.

Earlier coverage of this dataset reported high Gini coefficients but reported them at
the aggregate level, leaving open a question central to policy design: is the extreme
concentration a *micro-enterprise pathology* — i.e., a problem specific to the smallest
firms — or a *universal property* of the Brazilian formal sector?

The answer matters. If concentration is concentrated in micros, raising thresholds or
splitting micros into more granular tiers could bring meaningful redistribution. If
concentration is universal, threshold reform is cosmetic — the same power-law sits
behind every regulatory label, and revenue is the wrong targeting dimension.

## Research Question

> **Q1.** Is capital concentration in Brazilian formal-sector firms a feature of the
> micro-enterprise category specifically, or is the same power-law signature visible
> across every size tier?

## Design Conventions

```{r}
#| label: setup-theme

library(tidyverse)
library(scales)
library(patchwork)
library(boot)        # bootstrap Gini CIs
library(gt)          # publication-quality tables

# ── Shared palette ──────────────────────────────────────────────────────────
pal_size <- c(
  "Micro"        = "#c3553a",   # terracotta
  "Small"        = "#d9a441",   # amber
  "Medium/Large" = "#2a5a6b"    # deep teal
)

pal_segment <- c(
  "Top 1%"     = "#1d3557",
  "Next 9%"    = "#457b9d",
  "Next 40%"   = "#a8dadc",
  "Bottom 50%" = "#e9ecef"
)

# ── Shared theme ────────────────────────────────────────────────────────────
theme_cwc <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title        = element_text(face = "bold", size = base_size + 2,
                                       colour = "#1d3557", margin = margin(b = 4)),
      plot.subtitle     = element_text(colour = "grey35", size = base_size - 2,
                                       lineheight = 1.35, margin = margin(b = 18)),
      plot.caption      = element_text(colour = "grey55", size = base_size - 5,
                                       hjust = 0, margin = margin(t = 12)),
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      strip.text        = element_text(face = "bold", size = base_size - 1,
                                       colour = "#1d3557"),
      strip.background  = element_rect(fill = "grey96", colour = NA),
      panel.grid.minor  = element_blank(),
      panel.grid.major  = element_line(colour = "grey92", linewidth = 0.3),
      panel.spacing     = unit(1.4, "lines"),
      legend.position   = "top",
      legend.title      = element_text(face = "bold", size = base_size - 2),
      legend.text       = element_text(size = base_size - 2),
      axis.title        = element_text(colour = "grey30", size = base_size - 2),
      axis.text         = element_text(colour = "grey40", size = base_size - 3)
    )
}

caption_text <- paste(
  "Source: Brazilian CNPJ Registry · TidyTuesday 2026 Wk 4 (Extended)",
  "· Boyd, He & Miao · 2026-02-25"
)
```

## Data Preparation

```{r}
#| label: data-prep

companies <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv",
  show_col_types = FALSE
)

co <- companies |>
  filter(
    !is.na(company_size),
    !is.na(capital_stock),
    !is.na(owner_qualification),
    capital_stock > 0
  ) |>
  mutate(
    size = case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE                               ~ "Medium/Large"
    ),
    size = factor(size, levels = c("Micro", "Small", "Medium/Large")),
    log_capital = log10(capital_stock)
  )

cat("Observations after cleaning:", comma(nrow(co)), "\n")
cat("Tier counts:\n"); print(count(co, size))
```

## Descriptive Snapshot

Before any inequality measure, a sanity check on the raw distribution by tier. Means
exceed medians by an order of magnitude in every tier — the first hint that the
distribution is not just skewed but heavy-tailed.

```{r}
#| label: tbl-descriptive

co |>
  group_by(size) |>
  summarise(
    n        = n(),
    mean_cap = mean(capital_stock),
    p50      = median(capital_stock),
    p75      = quantile(capital_stock, 0.75),
    p90      = quantile(capital_stock, 0.90),
    p99      = quantile(capital_stock, 0.99),
    max_cap  = max(capital_stock),
    .groups  = "drop"
  ) |>
  gt() |>
  tab_header(
    title    = md("**Capital stock by firm size tier**"),
    subtitle = "All values in BRL · means dwarf medians in every tier"
  ) |>
  cols_label(
    size = "Tier", n = "N",
    mean_cap = "Mean", p50 = "Median",
    p75 = "P75", p90 = "P90", p99 = "P99",
    max_cap = "Max"
  ) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = c(mean_cap, p50, p75, p90, p99, max_cap),
             suffixing = TRUE, decimals = 1) |>
  tab_options(
    table.font.size            = px(13),
    heading.align              = "left",
    row_group.background.color = "#f4f4f4"
  ) |>
  tab_source_note(md("*Source: Brazilian CNPJ Registry*"))
```

## Analysis: Concentration is Universal Across Tiers

### Methodological motivation

Plain Lorenz curves are a poor visualization tool when inequality is extreme. When the
top 1% of firms holds 96–99% of capital, three Lorenz curves at Gini 0.98, 0.99, 0.995
plot as nearly identical near-L shapes — visual indistinguishability where the
analytically interesting separation actually lives. Two adjustments make the structure
readable:

1. **Discrete top-fraction ladder.** Instead of drawing the full Lorenz curve through
   every firm, we evaluate the cumulative share of capital at seven specific
   top-fraction breakpoints (Top 0.1%, 0.5%, 1%, 5%, 10%, 25%, 50%) and connect them
   on a log-scaled x-axis. This avoids two artefacts of full-data Lorenz curves —
   visual aliasing from millions of overplotted points, and the optical illusion of a
   "rapid rise" near the top that's actually just the geometry of an L. Each labelled
   point is a directly readable fact: "the top 1% of Micro firms holds 96% of Micro
   capital."
2. **Stacked rank-bucket decomposition.** Convert the Gini coefficient into a directly
   readable percentage by reporting the share of each tier's capital held by the top
   1%, next 9%, next 40%, and bottom 50%. This is harder to misread than Gini and
   communicates immediately to non-technical audiences.

We also report **bootstrap 95% confidence intervals on the Gini point estimates**
(B = 1,000 resamples per tier) to confirm that the small numerical differences between
tier Ginis are not artefacts of sample variance. With N at this scale, the CIs are
mechanically tight — but reporting them is good discipline.

### Bootstrap Gini estimation

```{r}
#| label: gini-bootstrap

gini <- function(x) {
  x <- sort(x[!is.na(x) & x > 0])
  n <- length(x)
  if (n < 2L) return(NA_real_)
  2 * sum(seq_len(n) * x) / (n * sum(x)) - (n + 1) / n
}

gini_stat <- function(data, i) gini(data[i])

set.seed(2026)
gini_ci <- co |>
  group_by(size) |>
  group_modify(~ {
    b  <- boot(.x$capital_stock, gini_stat, R = 1000)
    ci <- boot.ci(b, type = "perc", conf = 0.95)
    tibble(
      g     = b$t0,
      lower = ci$percent[4L],
      upper = ci$percent[5L],
      n     = nrow(.x)
    )
  }) |>
  ungroup()

# Top-1% share — robust complement to Gini in extreme-tail regimes
top1_share <- co |>
  group_by(size) |>
  arrange(desc(capital_stock), .by_group = TRUE) |>
  summarise(
    top1 = sum(capital_stock[seq_len(ceiling(n() * 0.01))]) / sum(capital_stock),
    .groups = "drop"
  )

gini_ci |>
  left_join(top1_share, by = "size") |>
  gt() |>
  tab_header(
    title    = md("**Bootstrapped Gini coefficients and top-1% capital shares**"),
    subtitle = "B = 1,000 percentile bootstrap resamples per tier"
  ) |>
  cols_label(
    size = "Tier", g = "Gini",
    lower = "CI Lower", upper = "CI Upper",
    n = "N", top1 = "Top 1% share"
  ) |>
  fmt_number(columns = c(g, lower, upper), decimals = 4) |>
  fmt_percent(columns = top1, decimals = 1) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0) |>
  tab_options(table.font.size = px(13), heading.align = "left")
```

### Two-panel concentration figure

```{r}
#| label: fig-concentration
#| fig-width: 12
#| fig-height: 6.0
#| fig-cap: "Capital concentration is near-total in every Brazilian firm-size tier."

# ── Panel A data: cumulative share at discrete top-fraction breakpoints ────
top_fracs  <- c(0.001, 0.005, 0.01, 0.05, 0.10, 0.25, 0.50)
top_labels <- c("Top\n0.1%", "Top\n0.5%", "Top\n1%", "Top\n5%",
                "Top\n10%", "Top\n25%", "Top\n50%")

cum_share_at <- function(x, f) {
  x_sorted <- sort(x, decreasing = TRUE)
  k <- max(1L, ceiling(length(x_sorted) * f))
  sum(x_sorted[seq_len(k)]) / sum(x_sorted)
}

ladder_data <- co |>
  group_by(size) |>
  group_modify(~ tibble(
    top_frac  = top_fracs,
    cum_share = vapply(top_fracs,
                       function(f) cum_share_at(.x$capital_stock, f),
                       numeric(1))
  )) |>
  ungroup()

# Legend labels: bootstrap-confirmed Gini, in the order Micro / Small / Med-Large
gini_in_order <- gini_ci$g[match(c("Micro", "Small", "Medium/Large"),
                                  as.character(gini_ci$size))]
legend_labels <- sprintf("%s — Gini %.3f",
                         c("Micro", "Small", "Medium/Large"),
                         gini_in_order)

# ── Panel A: discrete ladder concentration chart ───────────────────────────
p_ladder <- ggplot(ladder_data,
                   aes(x = top_frac, y = cum_share,
                       colour = size, group = size)) +
  geom_line(linewidth = 1.4, alpha = 0.9) +
  geom_point(size = 4.5, shape = 21, fill = "white", stroke = 1.8) +
  # Direct labels at the two anchor points readers care about
  geom_text(
    data = filter(ladder_data, top_frac %in% c(0.001, 0.01)),
    aes(label = scales::percent(cum_share, accuracy = 1)),
    vjust = -1.6, fontface = "bold", size = 3.4,
    show.legend = FALSE
  ) +
  scale_x_log10(
    breaks = top_fracs,
    labels = top_labels,
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0.55, 1.07),
    breaks = c(0.6, 0.7, 0.8, 0.9, 1.0),
    expand = c(0, 0),
    name   = "Cumulative share of tier's capital"
  ) +
  scale_colour_manual(
    values = pal_size,
    labels = legend_labels,
    name   = NULL
  ) +
  labs(
    x = "Top fraction of firms (log scale)",
    subtitle = "(A)  Even the top 0.1% holds the majority of every tier's capital\n        Three near-identical curves — concentration is universal across tiers"
  ) +
  theme_cwc() +
  theme(
    legend.position    = "top",
    plot.subtitle      = element_text(colour = "grey35", size = 11,
                                      lineheight = 1.3, margin = margin(b = 10)),
    panel.grid.major.x = element_line(colour = "grey94", linewidth = 0.25),
    axis.text.x        = element_text(size = 9, lineheight = 0.9)
  )

# ── Panel B data: stacked rank-bucket decomposition ────────────────────────
decomp <- co |>
  group_by(size) |>
  arrange(desc(capital_stock), .by_group = TRUE) |>
  mutate(
    rank_pct = row_number() / n(),
    segment  = case_when(
      rank_pct <= 0.01 ~ "Top 1%",
      rank_pct <= 0.10 ~ "Next 9%",
      rank_pct <= 0.50 ~ "Next 40%",
      TRUE             ~ "Bottom 50%"
    ),
    segment  = factor(segment,
                      levels = c("Bottom 50%", "Next 40%", "Next 9%", "Top 1%"))
  ) |>
  group_by(size, segment) |>
  summarise(cap = sum(capital_stock), .groups = "drop") |>
  group_by(size) |>
  mutate(share = cap / sum(cap)) |>
  ungroup()
scale_y_continuous(
  labels = percent_format(),
  limits = c(0.5, 1.02),     # 改这里：从 0.6 → 0.5
  expand = c(0, 0),
  name   = "Cumulative share of tier's capital"
)

# ── Panel B: stacked decomposition with clean Top-1% callout ──────────────
p_decomp <- ggplot(decomp, aes(x = size, y = share, fill = segment)) +
  geom_col(width = 0.70, colour = "white", linewidth = 1.5) +
  # Big bold percentage centered in the Top 1% segment
  geom_text(
    data = filter(decomp, segment == "Top 1%"),
    aes(label = percent(share, accuracy = 1)),
    position    = position_stack(vjust = 0.5),
    colour      = "white", fontface = "bold",
    size        = 9,
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = pal_segment,
    breaks = c("Top 1%", "Next 9%", "Next 40%", "Bottom 50%"),
    name   = NULL
  ) +
  scale_y_continuous(
    labels = percent_format(),
    expand = c(0, 0),
    limits = c(0, 1.02),
    name   = "Share of tier's total capital"
  ) +
  labs(
    x = NULL,
    subtitle = "(B)  Top 1% holds 96–99% in every tier\n        The remaining 99% of firms is barely a sliver"
  ) +
  theme_cwc() +
  theme(
    legend.position    = "top",
    panel.grid.major.x = element_blank(),
    plot.subtitle      = element_text(colour = "grey35", size = 11,
                                      lineheight = 1.3, margin = margin(b = 10))
  )


# ── Combine ────────────────────────────────────────────────────────────────
(p_ladder | p_decomp) +
  plot_layout(widths = c(1.55, 1)) +
  plot_annotation(
    title    = "Capital concentration is near-total in every Brazilian firm-size tier",
    subtitle = paste0(
      "All three tiers exhibit Gini > 0.98 with the top 1% holding 96–99% of capital. ",
      "'Micro-enterprise' is not uniquely unequal —\n",
      "the same power-law structure sits behind every regulatory category."
    ),
    caption  = caption_text,
    theme    = theme_cwc()
  ) &
  theme(plot.title = element_text(face = "bold", size = 16, colour = "#1d3557"))
```
### Interpretation

Three findings emerge.

**Even the top 0.1% already holds the majority.** Panel A drills below
Panel B's headline. The "top 1% holds 96–99%" framing is itself an
understatement — even the top 0.1% of firms (one in a thousand) already
holds 87–96% of each tier's capital. Roughly nine-tenths of the top-1%
concentration is delivered by the top 0.1% alone. Past the Top 1% mark the
curves are flat at ~99%: the remaining 99% of firms is collectively a
sliver. The tier ordering at the apex (Medium/Large 96% > Micro 87%) is
partly an N effect — Medium/Large's "top 0.1%" is a numerically tighter set
of firms — and should not be read as "Medium/Large is more unequal than
Micro." What the visual confirms is universality of *shape*: the same
logarithmic-rise-to-plateau structure sits behind every regulatory tier.

**The Ginis are statistically distinguishable but economically negligible.**
All three tier Ginis exceed 0.98, with bootstrap 95% CIs that are tight and
non-overlapping. Statistically the tiers differ; substantively all three
are pinned against the ceiling of the Gini scale. No reasonable inequality
target distinguishes 0.985 from 0.992.

**Top-1% capital share is the dominant fact.** In each tier, the top one
percent of firms holds 96–99% of that tier's capital. The bottom 50% holds
less than one percent. This is not "high inequality" in the colloquial
sense; it is a power-law in which nearly all economic mass sits at the top
of the distribution. Panel B converts this into a directly readable
percentage: in every tier, the top 1% bar is essentially the entire bar.

The "capital trap" narrative attached to micro-enterprises in earlier
framings is misplaced. The concentration is **structural to the entire
Brazilian formal sector**, not specific to the micro tier. This reframes
the analytical agenda: if the inequality *shape* is universal, the
substantive question is no longer *which tier is unequal* but *who occupies
the top of each tier* — and that question cannot be answered by size
classification. It requires governance information.

## Methodological Notes

- **Why a discrete ladder rather than a full Lorenz curve.** A continuous Lorenz or
  concentration curve drawn through every firm (N > 100,000 per tier) introduces
  rendering artefacts: anti-aliasing on dense overplotted lines and visual aliasing
  near the steep top end. The seven-point ladder is computationally identical at the
  evaluated points (cumulative share at top f% is exactly the cumsum at the f-quantile
  of the sorted distribution) but immune to those artefacts and exactly labellable.
  The two visual claims — that the curves nearly coincide, and that each labelled
  anchor is precise — are easier to defend on a ladder than on a continuous curve
  with millions of plotted segments.
- **Gini stability at extreme concentration.** When the top firms dominate, Gini
  becomes mechanically close to 1 and economically meaningful changes are compressed
  into the third decimal. Top-1% share is more interpretable in this regime and
  invariant to the lower-tail definition.
- **Bootstrap sample size.** With N > 100,000 per tier, B = 1,000 percentile
  bootstrap resamples produce CIs in the fourth decimal place. We report them as a
  discipline against assuming a single resampling could change the qualitative
  finding, not because they add new information.
- **Excluding zero / missing capital.** Reported zero capital often reflects
  registry-coding artefacts (entities awaiting capital declaration after
  registration) rather than genuinely zero-capital operating firms. Including zeros
  would inflate the bottom-share estimates artifactually.
- **Why not Theil decomposition?** A Theil-T within-vs-between decomposition is a
  natural next step, and earlier drafts of this analysis included one. We dropped
  it here because the universal-power-law finding makes the *between-tier* component
  uninformative: when every tier has Gini > 0.98, decomposing tells us only that
  within-tier inequality dominates, which we already know visually. The decomposition
  is a useful exercise when between-tier differences carry information; here, they
  don't.

## Forward Link

If concentration is universal but the question is *who sits at the top of each tier*,
the analytical next step is to ask whether **governance archetype** — the legal role
of the controlling person — predicts capital independently of size. Part II takes
that question, examining whether governance shifts the *level* of capital, the
*variance* of capital, or both.
