---
title: "Member 2 Zhijun He"
---

## Data and Setup

### Packages

```{r}
#| label: setup

library(tidyverse)
library(scales)
library(ggridges)
library(patchwork)
library(boot)          # bootstrap Gini confidence intervals
library(quantreg)      # quantile regression
library(broom)         # tidy model outputs
library(gt)            # publication-quality tables
library(modelsummary)  # regression summary tables
library(ggtext)        # markdown / HTML in ggplot labels
library(glue)          # string interpolation

# ── Shared palette ─────────────────────────────────────────────────────────────
pal_size <- c("Micro" = "#e76f51", "Small" = "#f4a261", "Medium/Large" = "#264653")
pal_gov  <- c(
  "Sole Entrepreneur"                    = "#e9c46a",
  "Managing Partner /\nAdministrator"    = "#4a90d9",
  "Beneficial Owner\n(passive/investor)" = "#264653"
)

caption_base <- "Source: Brazilian CNPJ Registry ·He, Miao & Boyd"
```

### Data Ingestion and Cleaning

```{r}
#| label: data-load

companies <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv",
  show_col_types = FALSE
)

# ── Core recoding ───────────────────────────────────────────────────────────────
co <- companies |>
  # Drop records with missing or non-positive capital (not usable for log-scale analysis)
  filter(
    !is.na(company_size),
    !is.na(capital_stock),
    capital_stock > 0,
    !is.na(owner_qualification)
  ) |>
  mutate(
    # Size tier (ordered factor)
    size = case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE                               ~ "Medium/Large"
    ),
    size = factor(size, levels = c("Micro", "Small", "Medium/Large")),

    # Governance archetype: collapse owner_qualification into 4 classes
    governance = case_when(
      str_detect(owner_qualification,
                 regex("beneficial owner", ignore_case = TRUE)) ~
        "Beneficial Owner\n(passive/investor)",
      str_detect(owner_qualification,
                 regex("entrepreneur|business owner", ignore_case = TRUE)) ~
        "Sole Entrepreneur",
      str_detect(owner_qualification,
                 regex("managing partner|administrator", ignore_case = TRUE)) ~
        "Managing Partner /\nAdministrator",
      TRUE ~ "Other"
    ),
    governance = factor(governance, levels = c(
      "Sole Entrepreneur",
      "Managing Partner /\nAdministrator",
      "Beneficial Owner\n(passive/investor)",
      "Other"
    )),

    # Log10 capital for distributional analysis
    log_capital = log10(capital_stock),

    # Legal-form shorthand for micro analysis
    legal_short = case_when(
      str_detect(legal_nature, "LLC")                 ~ "LLC",
      str_detect(legal_nature, "Sole Proprietorship") ~ "Sole Prop.",
      str_detect(legal_nature, "Individual Limited")  ~ "EIRELI",
      str_detect(legal_nature, "Simple")              ~ "Simple Part.",
      str_detect(legal_nature, "Silent")              ~ "Silent Part.",
      str_detect(legal_nature, "Corporation")         ~ "Corporation",
      TRUE                                            ~ "Other"
    ),

    # Capital-rich flag: top quartile *within* micro-enterprises
    capital_q75_micro     = quantile(capital_stock[size == "Micro"], 0.75, na.rm = TRUE),
    is_capital_rich_micro = (size == "Micro") & (capital_stock >= capital_q75_micro)
  )

cat("Final observations:", scales::comma(nrow(co)), "\n")
```

***

## Analysis II — The Governance Premium {#sec-governance}

### Capital Distribution by Governance Archetype

We examine the unconditional distribution of log-capital by governance archetype within each
size tier. Quantile lines at Q1, Q2, and Q3 are shown. Kruskal-Wallis tests (non-parametric,
appropriate given non-normal capital distributions) evaluate whether governance predicts capital
within each tier.

```{r}
#| label: governance-ridges
#| fig-width: 11
#| fig-height: 6.5

# Kruskal-Wallis p-values per size tier
kw_pvals <- co |>
  filter(governance != "Other") |>
  group_by(size) |>
  summarise(
    p = kruskal.test(capital_stock ~ governance)$p.value,
    .groups = "drop"
  ) |>
  mutate(
    p_lab = if_else(p < 0.001, "KW: p < 0.001", glue("KW: p = {round(p, 3)}"))
  )

gov_medians <- co |>
  filter(governance != "Other") |>
  group_by(size, governance) |>
  summarise(med_cap = median(capital_stock), n = n(), .groups = "drop") |>
  mutate(
    med_lab = glue("md: {number(med_cap, scale = 1e-3, suffix = 'k', accuracy = 1)}")
  )

ggplot(
  co |> filter(governance != "Other"),
  aes(x = log_capital, y = governance, fill = governance)
) +
  geom_density_ridges(
    alpha          = 0.75, scale = 0.88,
    quantile_lines = TRUE, quantiles = c(0.25, 0.5, 0.75),
    colour = "white", linewidth = 0.35
  ) +
  geom_text(
    data        = gov_medians,
    aes(x = log10(med_cap) + 0.05, y = governance, label = med_lab),
    inherit.aes = FALSE,
    size = 2.8, colour = "grey20", hjust = 0, vjust = -1.5
  ) +
  geom_text(
    data        = kw_pvals,
    aes(x = Inf, y = Inf, label = p_lab),
    inherit.aes = FALSE,
    hjust = 1.1, vjust = 1.5,
    size = 3.0, colour = "grey30", fontface = "italic"
  ) +
  facet_wrap(~size, nrow = 1) +
  scale_fill_manual(values = pal_gov, guide = "none") +
  scale_x_continuous(
    breaks = c(5, 5.5, 6, 6.5, 7, 7.5, 8),
    labels = c("100k", "316k", "1M", "3.2M", "10M", "32M", "100M"),
    name   = "Capital stock (BRL, log10 scale)"
  ) +
  labs(
    title    = "Beneficial Owners command higher capital at **every** size tier",
    subtitle = paste0(
      "Density ridges by governance archetype within each firm-size tier.\n",
      "Vertical white lines mark Q1, median, and Q3. ",
      "KW = Kruskal-Wallis non-parametric group comparison."
    ),
    y       = NULL,
    caption = caption_base
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_markdown(face = "bold", size = 14, colour = "#264653"),
    plot.subtitle      = element_text(colour = "grey40", size = 10.5,
                                      lineheight = 1.3, margin = margin(b = 14)),
    plot.caption       = element_text(colour = "grey50", size = 8, hjust = 0),
    strip.text         = element_text(face = "bold", size = 12, colour = "#264653"),
    strip.background   = element_rect(fill = "grey95", colour = NA),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

### Quantile Regression: Does the Governance Premium Vary Across the Distribution?

A median regression (τ = 0.5) describes the typical firm, but development policy is most
concerned with the *lower* tail (genuinely capital-constrained entrepreneurs) and the *upper*
tail (well-capitalized investment vehicles masquerading as micro-enterprises). Quantile regression
at τ ∈ {0.10, 0.25, 0.50, 0.75, 0.90} reveals whether the governance premium is uniform or
systematically concentrated at the extremes.

```{r}
#| label: quantile-regression
#| fig-width: 11
#| fig-height: 5.5
#| cache: true

taus <- c(0.10, 0.25, 0.50, 0.75, 0.90)

# Recode to short labels — keeps coefficient names clean
qr_data <- co |>
  filter(governance != "Other") |>
  mutate(
    gov = fct_recode(governance,
      "Managing Partner" = "Managing Partner /\nAdministrator",
      "Beneficial Owner" = "Beneficial Owner\n(passive/investor)"
    ),
    gov = relevel(gov, ref = "Sole Entrepreneur")
  )

# Point estimates only — avoids singular matrix from NID SE
qr_tidy <- map_dfr(taus, \(tau) {
  fit <- rq(log_capital ~ gov + size, tau = tau, data = qr_data, method = "fn")
  tibble(term = names(coef(fit)), estimate = coef(fit), tau = tau)
}) |>
  filter(str_detect(term, "^gov")) |>
  mutate(gov_label    = str_remove(term, "^gov"),
         effect_ratio = 10^estimate)

# OLS mean-effect comparison
ols_coef <- lm(log_capital ~ gov + size, data = qr_data) |>
  tidy() |>
  filter(str_detect(term, "^gov")) |>
  mutate(gov_label    = str_remove(term, "^gov"),
         effect_ratio = 10^estimate)

pal_gov_short <- c("Managing Partner" = "#4a90d9", "Beneficial Owner" = "#264653")

ggplot(qr_tidy, aes(x = tau, y = effect_ratio, colour = gov_label)) +
  geom_hline(yintercept = 1, colour = "grey60",
             linetype = "dotted", linewidth = 0.5) +
  geom_hline(data = ols_coef,
             aes(yintercept = effect_ratio, colour = gov_label),
             linetype = "dashed", linewidth = 0.65, alpha = 0.55) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  facet_wrap(~gov_label, nrow = 1, scales = "free_y") +
  scale_colour_manual(values = pal_gov_short, guide = "none") +
  scale_x_continuous(breaks = taus, labels = paste0("t=", taus)) +
  scale_y_continuous(labels = \(x) paste0(round(x, 2), "x"),
                     name   = "Capital multiplier vs. Sole Entrepreneur baseline") +
  labs(
    title    = "The governance premium **widens at the upper tail** — especially for Beneficial Owners",
    subtitle = paste0(
      "Quantile regression point estimates (exponentiated) at tau = 0.10, 0.25, 0.50, 0.75, 0.90.\n",
      "Baseline = Sole Entrepreneur. Dashed line = OLS conditional mean."
    ),
    x = "Quantile (tau)", caption = caption_base
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_markdown(face = "bold", size = 14, colour = "#264653"),
    plot.subtitle    = element_text(colour = "grey40", size = 10.2,
                                    lineheight = 1.3, margin = margin(b = 14)),
    plot.caption     = element_text(colour = "grey50", size = 8, hjust = 0),
    strip.text       = element_text(face = "bold", size = 11, colour = "#264653"),
    strip.background = element_rect(fill = "grey95", colour = NA),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 9)
  )
```

The quantile regression reveals a pattern OLS obscures: the governance premium is not uniform.
For Beneficial Owners, the capital multiplier relative to sole entrepreneurs grows substantially
from the 10th to the 90th percentile. This asymmetry is diagnostic — Beneficial Owner-controlled
firms are not simply "shifted up" relative to sole entrepreneurs but are a structurally different
population that saturates the upper tail of the capital distribution. The OLS estimate (dashed
line) understates the premium at the upper tail and overstates it at the lower tail.

---
