---
title: "Capital Without Category — Part III"
subtitle: "The partnership channel: legal form × governance and the micro-enterprise escape route"
author: "Paxton Boyd"
date: "2026-02-25"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-dpi: 180
    fig-align: center
    fig-cap-location: bottom
    df-print: paged
execute:
  warning: false
  message: false
  cache: true
---

## Background

Parts I and II established two facts. **First**, capital concentration in Brazilian
formal-sector firms is universal across regulatory tiers — every firm-size category
has Gini > 0.98 and a top-1% capital share above 96%. **Second**, governance
archetype shapes the capital distribution in distributional terms: Managing Partners
lead the median (and scale with size), Sole Entrepreneurs stay flat, and Beneficial
Owners carry a heavy upper tail at every size.

The combined picture leaves a sharp residual question. Inside the micro-enterprise
tier — where Brazil's main small-business support programs are targeted — there is
a non-trivial subpopulation of firms with capital stock far above what their revenue
classification predicts. These are firms whose annual gross revenue is below the
360,000 BRL Simples Nacional threshold but whose declared capital is above the
75th percentile of all micros. They are *capital-rich micros*: they wear the
regulatory clothing of a small business but hold the capital stock of something
larger.

If the goal of micro-enterprise policy is to support genuinely capital-constrained
operators, capital-rich micros are a targeting failure: any subsidy or simplified-
tax treatment they receive flows to firms that did not need it. The empirical
question is what *legal-structural signature* these firms share. If a single legal
form (say, EIRELI — Brazil's individual limited-liability vehicle) accounts for
most of them, the policy fix is straightforward: condition support on legal form
in addition to revenue. If the signature is more textured — a specific
combination of legal form *and* governance archetype — the fix has to be designed
around that combination.

## Research Question

> **Q3.** Among micro-enterprises, which combinations of legal form and governance
> archetype produce the highest share of capital-rich firms? Is the escape route a
> single legal form (the popular "EIRELI" framing), or a structural combination
> that no single dimension captures?

We define a **capital-rich micro** as a micro-enterprise whose capital stock falls
in the top quartile of all micros' capital — i.e., above the 75th percentile,
which in this dataset is approximately BRL 500,000. This threshold is descriptive,
not normative; it captures the upper tail of the micro distribution where
"escape from the typical micro profile" begins to be analytically meaningful.

## Design Conventions

```{r}
#| label: setup-theme

library(tidyverse)
library(scales)
library(gt)

# ── Shared palette ──────────────────────────────────────────────────────────
pal_size <- c(
  "Micro"        = "#c3553a",
  "Small"        = "#d9a441",
  "Medium/Large" = "#2a5a6b"
)

pal_gov <- c(
  "Sole Entrepreneur"                    = "#d9a441",
  "Managing Partner /\nAdministrator"    = "#5b8a9a",
  "Beneficial Owner\n(passive/investor)" = "#1d3557"
)

# ── Shared theme ────────────────────────────────────────────────────────────
theme_cwc <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title        = element_text(face = "bold", size = base_size + 2,
                                       colour = "#1d3557", margin = margin(b = 4)),
      plot.subtitle     = element_text(colour = "grey35", size = base_size - 2,
                                       lineheight = 1.35, margin = margin(b = 18)),
      plot.caption      = element_text(colour = "grey55", size = base_size - 5,
                                       hjust = 0, margin = margin(t = 12)),
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      strip.text        = element_text(face = "bold", size = base_size - 1,
                                       colour = "#1d3557"),
      strip.background  = element_rect(fill = "grey96", colour = NA),
      panel.grid.minor  = element_blank(),
      panel.grid.major  = element_line(colour = "grey92", linewidth = 0.3),
      panel.spacing     = unit(1.4, "lines"),
      legend.position   = "top",
      legend.title      = element_text(face = "bold", size = base_size - 2),
      legend.text       = element_text(size = base_size - 2),
      axis.title        = element_text(colour = "grey30", size = base_size - 2),
      axis.text         = element_text(colour = "grey40", size = base_size - 3)
    )
}

caption_text <- paste(
  "Source: Brazilian CNPJ Registry · TidyTuesday 2026 Wk 4 (Extended)",
  "· Boyd, He & Miao · 2026-02-25"
)
```

## Data Preparation

```{r}
#| label: data-prep

companies <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv",
  show_col_types = FALSE
)

co <- companies |>
  filter(
    !is.na(company_size),
    !is.na(capital_stock),
    !is.na(owner_qualification),
    capital_stock > 0
  ) |>
  mutate(
    size = case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE                               ~ "Medium/Large"
    ),
    size = factor(size, levels = c("Micro", "Small", "Medium/Large")),

    governance = case_when(
      str_detect(owner_qualification,
                 regex("beneficial owner", ignore_case = TRUE))     ~ "Beneficial Owner\n(passive/investor)",
      str_detect(owner_qualification,
                 regex("entrepreneur|business owner", ignore_case = TRUE)) ~ "Sole Entrepreneur",
      str_detect(owner_qualification,
                 regex("managing partner|administrator", ignore_case = TRUE)) ~ "Managing Partner /\nAdministrator",
      TRUE ~ "Other"
    ),
    governance = factor(governance, levels = c(
      "Sole Entrepreneur",
      "Managing Partner /\nAdministrator",
      "Beneficial Owner\n(passive/investor)",
      "Other"
    )),

    log_capital = log10(capital_stock),

    # Capital-rich threshold = 75th percentile within micros
    capital_q75_micro     = quantile(capital_stock[size == "Micro"], 0.75, na.rm = TRUE),
    is_capital_rich_micro = (size == "Micro") & (capital_stock >= capital_q75_micro)
  )

cat("Observations:", comma(nrow(co)), "\n")
cat("Micros total:", comma(sum(co$size == "Micro")), "\n")
cat("Capital-rich threshold: BRL ",
    comma(unique(co$capital_q75_micro[!is.na(co$capital_q75_micro)])[1]),
    "\n", sep = "")
```

## Analysis: The Escape Route is Partnership-with-Passive-Equity

### Methodological motivation

The simplest analytical approach — comparing escape rates by legal form alone —
would obscure the answer, because legal form and governance are correlated but not
identical. A Sole Proprietorship is mechanically tied to a Sole Entrepreneur (one
operator, one identity), but an LLC or a Simple Partnership can be controlled by
either an active Managing Partner or a passive Beneficial Owner. Collapsing across
governance averages over these very different ownership mechanics.

We therefore compute escape rates at the **legal-form × governance cell level**
within the micro tier. Two methodological details support credibility:

1. **Cell-size threshold.** We exclude any cell with fewer than 20 observations.
   With small n, sample-proportion estimates are noisy, and the visualization
   would be dominated by extreme rates from cells of n = 3 or n = 5. The cutoff is
   conservative; a sensitivity check at n ≥ 50 produces the same ranking.
2. **Wilson 95% confidence intervals.** For binomial proportions, the textbook
   normal-approximation (Wald) interval performs poorly when p̂ is near 0 or 1, or
   when n is moderate. The Wilson score interval has better coverage in these
   regimes and avoids the embarrassment of CIs that extend below 0 or above 1. We
   report Wilson intervals on every cell rate.

### Cell-level escape rates with Wilson intervals

```{r}
#| label: tbl-escape-rates

micro_profile <- co |>
  filter(size == "Micro", governance != "Other") |>
  mutate(
    legal_short = case_when(
      str_detect(legal_nature, "LLC")                 ~ "LLC",
      str_detect(legal_nature, "Sole Proprietorship") ~ "Sole Proprietary",
      str_detect(legal_nature, "Individual Limited")  ~ "EIRELI",
      str_detect(legal_nature, "Simple")              ~ "Simple Partnership",
      str_detect(legal_nature, "Silent")              ~ "Silent Partnership",
      str_detect(legal_nature, "Corporation")         ~ "Corporation",
      TRUE                                            ~ "Other"
    )
  )

# Wilson score interval for a binomial proportion
wilson_ci <- function(k, n, conf = 0.95) {
  z  <- qnorm(1 - (1 - conf) / 2)
  p  <- k / n
  lo <- (p + z^2 / (2 * n) -
           z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))) / (1 + z^2 / n)
  hi <- (p + z^2 / (2 * n) +
           z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))) / (1 + z^2 / n)
  tibble(ci_lo = lo, ci_hi = hi)
}

escape_rate <- micro_profile |>
  group_by(legal_short, governance) |>
  summarise(
    n_total    = n(),
    n_rich     = sum(is_capital_rich_micro),
    escape_pct = n_rich / n_total,
    .groups    = "drop"
  ) |>
  filter(n_total >= 20) |>
  mutate(
    wilson_ci(n_rich, n_total),                    # splices ci_lo, ci_hi
    combo_label = paste0(legal_short, " × ",
                         str_replace(governance, "\\n", " "))
  ) |>
  arrange(desc(escape_pct))

escape_rate |>
  select(legal_short, governance, n_total, n_rich,
         escape_pct, ci_lo, ci_hi) |>
  mutate(governance = str_replace(governance, "\n", " ")) |>
  gt() |>
  tab_header(
    title    = md("**Capital-rich micro escape rates by legal form × governance**"),
    subtitle = "Wilson 95% confidence intervals · cells with n < 20 excluded"
  ) |>
  cols_label(
    legal_short = "Legal form",
    governance  = "Governance archetype",
    n_total     = "N",
    n_rich      = "N rich",
    escape_pct  = "Escape %",
    ci_lo       = "CI Lower",
    ci_hi       = "CI Upper"
  ) |>
  fmt_number(columns = c(n_total, n_rich), use_seps = TRUE, decimals = 0) |>
  fmt_percent(columns = c(escape_pct, ci_lo, ci_hi), decimals = 1) |>
  tab_options(table.font.size = px(13), heading.align = "left")
```

### Independence test: are legal form and capital-rich status linked?

A formal omnibus check before the visual: pooling across governance, is the
legal-form distribution different among capital-rich micros than among
non-capital-rich micros? If the χ² test fails to reject, the cell-level pattern
might be noise. If it rejects, the cells are carrying real structure.

```{r}
#| label: chisq-test

chisq_tab <- micro_profile |>
  count(legal_short, is_capital_rich_micro) |>
  pivot_wider(names_from  = is_capital_rich_micro,
              values_from = n,
              values_fill = 0,
              names_prefix = "rich_") |>
  filter((rich_TRUE + rich_FALSE) >= 20) |>
  column_to_rownames("legal_short") |>
  as.matrix()

chisq_res <- chisq.test(chisq_tab)

cat("Pearson χ² test of legal form × capital-rich status (within micros):\n")
cat("  χ² statistic:", round(chisq_res$statistic, 1), "\n")
cat("  df:          ", chisq_res$parameter, "\n")
cat("  p-value:     ", format.pval(chisq_res$p.value, eps = 1e-300), "\n")
```

The null of independence is rejected decisively. Legal form and capital-rich
status are not independent within micros — the cell-level visualization that
follows is reading real signal.

### Visualization: the ranked escape profile

```{r}
#| label: fig-escape
#| fig-width: 11
#| fig-height: 6
#| fig-cap: "Capital-rich-micro share by legal form × governance combination."

ggplot(
  escape_rate,
  aes(x = fct_reorder(combo_label, escape_pct),
      y = escape_pct,
      fill = governance)
) +
  geom_col(width = 0.7, alpha = 0.95) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0.18, colour = "grey30", linewidth = 0.55) +
  geom_text(aes(label = percent(escape_pct, accuracy = 0.1),
                y = ci_hi),
            hjust = -0.18, size = 3.3, colour = "grey20") +
  geom_text(aes(label = paste0("n = ", comma(n_total)), y = 0.003),
            hjust = 0, size = 2.8, colour = "white", fontface = "bold") +
  coord_flip(clip = "off") +
  scale_y_continuous(
    labels = percent_format(),
    limits = c(0, 0.62),
    expand = expansion(mult = c(0, 0.05)),
    name   = "Share of cell above the 75th capital percentile (BRL 500,000)"
  ) +
  scale_fill_manual(values = pal_gov, name = "Governance archetype",
                    guide = guide_legend(reverse = TRUE)) +
  labs(
    title    = "The micro-enterprise escape profile is a partnership with a passive equity holder",
    subtitle = paste0(
      "Share of capital-rich micros (>BRL 500k) by legal form × governance cell. ",
      "Error bars = Wilson 95% CI.\n",
      "Cells with n < 20 excluded. Counts shown in white."
    ),
    x        = NULL,
    caption  = caption_text
  ) +
  theme_cwc() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y        = element_text(size = 10, colour = "grey20"),
    plot.margin        = margin(10, 40, 10, 10)
  )
```

### Interpretation

Ranking the cells that survive the n ≥ 20 cutoff produces a clean ordering.

**Simple Partnership × Beneficial Owner sits at the top (~48%).** Almost half of
the micros in this cell are capital-rich. The Wilson interval excludes any rate
below the cell-pooled mean; the result is not a small-sample artifact.

**Simple Partnership × Managing Partner is second (~42%).** The same legal form,
with an active rather than passive controlling party, still produces well-above-
average escape rates. The legal form itself is doing work, independent of who
controls it.

**Sole Proprietary × Sole Entrepreneur, LLC × Managing Partner, EIRELI ×
Beneficial Owner trail behind** (in the low-30s and below). EIRELI in particular
is *not* the dominant escape route — its rate is below the LLC × Managing Partner
cell. The popular "EIRELI as escape vehicle" framing is not what the data show.

The pattern is a **two-dimensional combination, not a one-dimensional channel**:

- **Partnership law is doing structural work.** The Simple Partnership is a
  multi-party form requiring two or more named parties on paper. Its presence at
  the top of the ranking — across both governance archetypes — suggests that the
  partnership-form requirement of multiple registered parties opens a channel
  for capital that single-person forms (EIRELI, Sole Proprietorship) cannot
  match, even when those single-person forms permit comparable nominal capital.
- **Passive ownership amplifies the channel.** Within the Simple Partnership
  form, switching from Managing Partner to Beneficial Owner control adds about
  six percentage points to the escape rate. The same legal architecture, more
  capital, when the controlling party is passive equity rather than active
  operator.

The combined signature — partnership law plus passive equity — is the fingerprint
of holding-company architecture operating through Brazil's small-business legal
infrastructure: a passive equity holder channels capital into a nominally
multi-party vehicle whose annual revenue happens to sit below the Simples
threshold. The vehicle is legally a micro-enterprise; structurally it is a
subsidiary.

## Methodological Notes

- **Why not logistic regression here.** An earlier draft estimated a logistic
  model regressing capital-rich status on governance and legal form with
  reference categories. The cell-level visualization communicates the same
  structural finding more directly because it makes the *interaction* between
  legal form and governance visible — which is the substantive result. A logistic
  model with main effects only would average across the interaction; a model
  with the full interaction would produce 12 coefficients no easier to read than
  the cells themselves.
- **Why the 75th percentile cutoff.** The threshold is descriptive, not
  normative. We tested 80th and 90th percentile cutoffs as sensitivity; the
  ranking of cells is stable, though absolute rates of course shift down with a
  stricter threshold.
- **What we cannot say.** This is a descriptive analysis of administrative
  registry data. We are not estimating a treatment effect. The capital-rich
  micros may be choosing partnership form *because* they have capital (selection
  on capacity) rather than acquiring capital because they chose partnership form
  (legal-form-as-channel). Distinguishing these requires panel data or a
  registration-reform discontinuity, neither of which the CNPJ snapshot
  supports.

## Synthesis: The Trilogy

Three findings link up into a single argument about the limits of revenue-based
firm classification.

**Part I — concentration is universal.** Capital concentration in Brazilian
formal-sector firms is a power-law structure shared by every size tier (Gini >
0.98 everywhere; top-1% share of 96–99%). Treating "micro-enterprise" as a
uniquely unequal category misreads the data. The same shape sits behind every
regulatory label.

**Part II — governance is distributional.** Governance archetype shapes the
capital distribution differently across types: Managing Partners deliver a level
premium that scales with size, Sole Entrepreneurs stay flat (consistent with the
agency-theory prediction for solo proprietors), and Beneficial Owners produce a
heavy-tail distribution where most firms are small and a minority hold enormous
capital. Size and governance are orthogonal dimensions of the registry, and
treating them as substitutes systematically misallocates.

**Part III — the escape route is structural.** Among micro-enterprises, the
capital-rich subpopulation concentrates in **Simple Partnership × Beneficial
Owner** and **Simple Partnership × Managing Partner** cells — partnership law
plus equity ownership rather than the popular EIRELI framing. This is the
fingerprint of holding-company architecture operating through small-business
legal infrastructure.

The policy implication is direct. If Brazil's development goal is to support
genuinely capital-constrained entrepreneurs, **revenue alone is a porous
targeting instrument**. The registry already carries the information needed to
do better: legal form and governance archetype are recorded for every entity at
registration, both available without survey or audit cost. A targeting rule
that conditions support on the *combination* — small revenue, single-party
legal form, sole or active governance — would exclude the partnership-with-
passive-equity escape route without imposing additional reporting burden on the
firms most policy is meant to reach.

The deeper structural point is the same one the analysis opened with.
Categories built on a single dimension (here, revenue) cannot do the work of
multi-dimensional structure (here, legal form × governance × revenue). When the
real heterogeneity is multi-dimensional and the targeting tool is one-
dimensional, the tool fails on its own terms — and the failure is not random
noise but a systematic bias toward the architecturally sophisticated.
