---
title: "Capital Without Category — Part I"
subtitle: "The universal power law: capital concentration across Brazilian firm-size tiers"
author: "Zhijun He"
date: "2026-02-25"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-dpi: 180
    fig-align: center
    fig-cap-location: bottom
    df-print: paged
execute:
  warning: false
  message: false
  cache: true
---

## Background

Brazil's official firm-size classification (Simples Nacional) is built on a single
threshold: **annual gross revenue**. A sole proprietor earning 360,000 BRL is legally a
"micro-enterprise" regardless of whether their declared capital stock is 50,000 BRL or
5,000,000 BRL. This regulatory choice creates a natural experiment in the CNPJ registry:
firms within the same legal category can have wildly different capital endowments.

Earlier coverage of this dataset reported high Gini coefficients but reported them at
the aggregate level, leaving open a question central to policy design: is the extreme
concentration a *micro-enterprise pathology* — i.e., a problem specific to the smallest
firms — or a *universal property* of the Brazilian formal sector?

The answer matters. If concentration is concentrated in micros, raising thresholds or
splitting micros into more granular tiers could bring meaningful redistribution. If
concentration is universal, threshold reform is cosmetic — the same power-law sits
behind every regulatory label, and revenue is the wrong targeting dimension.

## Research Question

> **Q1.** Is capital concentration in Brazilian formal-sector firms a feature of the
> micro-enterprise category specifically, or is the same power-law signature visible
> across every size tier?

## Design Conventions

```{r}
#| label: setup-theme

library(tidyverse)
library(scales)
library(patchwork)
library(boot)        # bootstrap Gini CIs
library(gt)          # publication-quality tables

# ── Shared palette ──────────────────────────────────────────────────────────
pal_size <- c(
  "Micro"        = "#c3553a",   # terracotta
  "Small"        = "#d9a441",   # amber
  "Medium/Large" = "#2a5a6b"    # deep teal
)

pal_segment <- c(
  "Top 1%"     = "#1d3557",
  "Next 9%"    = "#457b9d",
  "Next 40%"   = "#a8dadc",
  "Bottom 50%" = "#e9ecef"
)

# ── Shared theme ────────────────────────────────────────────────────────────
theme_cwc <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title        = element_text(face = "bold", size = base_size + 2,
                                       colour = "#1d3557", margin = margin(b = 4)),
      plot.subtitle     = element_text(colour = "grey35", size = base_size - 2,
                                       lineheight = 1.35, margin = margin(b = 18)),
      plot.caption      = element_text(colour = "grey55", size = base_size - 5,
                                       hjust = 0, margin = margin(t = 12)),
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      strip.text        = element_text(face = "bold", size = base_size - 1,
                                       colour = "#1d3557"),
      strip.background  = element_rect(fill = "grey96", colour = NA),
      panel.grid.minor  = element_blank(),
      panel.grid.major  = element_line(colour = "grey92", linewidth = 0.3),
      panel.spacing     = unit(1.4, "lines"),
      legend.position   = "top",
      legend.title      = element_text(face = "bold", size = base_size - 2),
      legend.text       = element_text(size = base_size - 2),
      axis.title        = element_text(colour = "grey30", size = base_size - 2),
      axis.text         = element_text(colour = "grey40", size = base_size - 3)
    )
}

caption_text <- paste(
  "Source: Brazilian CNPJ Registry · TidyTuesday 2026 Wk 4 (Extended)",
  "· Boyd, He & Miao · 2026-02-25"
)
```

## Data Preparation

```{r}
#| label: data-prep

companies <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv",
  show_col_types = FALSE
)

co <- companies |>
  filter(
    !is.na(company_size),
    !is.na(capital_stock),
    !is.na(owner_qualification),
    capital_stock > 0
  ) |>
  mutate(
    size = case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE                               ~ "Medium/Large"
    ),
    size = factor(size, levels = c("Micro", "Small", "Medium/Large")),
    log_capital = log10(capital_stock)
  )

cat("Observations after cleaning:", comma(nrow(co)), "\n")
cat("Tier counts:\n"); print(count(co, size))
```

## Descriptive Snapshot

Before any inequality measure, a sanity check on the raw distribution by tier. Means
exceed medians by an order of magnitude in every tier — the first hint that the
distribution is not just skewed but heavy-tailed.

```{r}
#| label: tbl-descriptive

co |>
  group_by(size) |>
  summarise(
    n        = n(),
    mean_cap = mean(capital_stock),
    p50      = median(capital_stock),
    p75      = quantile(capital_stock, 0.75),
    p90      = quantile(capital_stock, 0.90),
    p99      = quantile(capital_stock, 0.99),
    max_cap  = max(capital_stock),
    .groups  = "drop"
  ) |>
  gt() |>
  tab_header(
    title    = md("**Capital stock by firm size tier**"),
    subtitle = "All values in BRL · means dwarf medians in every tier"
  ) |>
  cols_label(
    size = "Tier", n = "N",
    mean_cap = "Mean", p50 = "Median",
    p75 = "P75", p90 = "P90", p99 = "P99",
    max_cap = "Max"
  ) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = c(mean_cap, p50, p75, p90, p99, max_cap),
             suffixing = TRUE, decimals = 1) |>
  tab_options(
    table.font.size            = px(13),
    heading.align              = "left",
    row_group.background.color = "#f4f4f4"
  ) |>
  tab_source_note(md("*Source: Brazilian CNPJ Registry*"))
```

## Analysis: Concentration is Universal Across Tiers

### Methodological motivation

Plain Lorenz curves are a poor visualization tool when inequality is extreme. When the
top 1% of firms holds something like 96–99% of capital, three Lorenz curves at Gini
0.98, 0.99, 0.995 plot as nearly identical near-L shapes — visual indistinguishability
where the analytically interesting separation actually lives. Two adjustments make the
structure readable:

1. **Log-rank concentration curves.** Plot cumulative share of capital against the
   *log* of rank percentile (top 0.1%, top 1%, top 10%, top 50%). Logarithmic rescaling
   on the rank axis expands the upper tail where the dispersion across tiers is
   concentrated. What looks like a single curve on linear scale resolves into separate
   curves once the top decile is given its proper visual weight.
2. **Stacked rank-bucket decomposition.** Convert the Gini coefficient into a directly
   readable percentage by reporting the share of each tier's capital held by the top
   1%, next 9%, next 40%, and bottom 50%. This is harder to misread than Gini and
   communicates immediately to non-technical audiences.

We also report **bootstrap 95% confidence intervals on the Gini point estimates**
(B = 1,000 resamples per tier) to confirm that the small numerical differences between
tier Ginis are not artifacts of sample variance. With N at this scale, the CIs are
mechanically tight — but reporting them is good discipline.

### Bootstrap Gini estimation

```{r}
#| label: gini-bootstrap

gini <- function(x) {
  x <- sort(x[!is.na(x) & x > 0])
  n <- length(x)
  if (n < 2L) return(NA_real_)
  2 * sum(seq_len(n) * x) / (n * sum(x)) - (n + 1) / n
}

gini_stat <- function(data, i) gini(data[i])

set.seed(2026)
gini_ci <- co |>
  group_by(size) |>
  group_modify(~ {
    b  <- boot(.x$capital_stock, gini_stat, R = 1000)
    ci <- boot.ci(b, type = "perc", conf = 0.95)
    tibble(
      g     = b$t0,
      lower = ci$percent[4L],
      upper = ci$percent[5L],
      n     = nrow(.x)
    )
  }) |>
  ungroup()

# Top-1% share — robust complement to Gini in extreme-tail regimes
top1_share <- co |>
  group_by(size) |>
  arrange(desc(capital_stock), .by_group = TRUE) |>
  summarise(
    top1 = sum(capital_stock[seq_len(ceiling(n() * 0.01))]) / sum(capital_stock),
    .groups = "drop"
  )

tier_stats <- gini_ci |>
  left_join(top1_share, by = "size") |>
  mutate(
    label = sprintf("%-13s  Gini %.3f  [%.3f, %.3f]   Top 1%% holds %s",
                    size, g, lower, upper,
                    percent(top1, accuracy = 1))
  )

gini_ci |>
  left_join(top1_share, by = "size") |>
  gt() |>
  tab_header(
    title    = md("**Bootstrapped Gini coefficients and top-1% capital shares**"),
    subtitle = "B = 1,000 percentile bootstrap resamples per tier"
  ) |>
  cols_label(
    size = "Tier", g = "Gini",
    lower = "CI Lower", upper = "CI Upper",
    n = "N", top1 = "Top 1% share"
  ) |>
  fmt_number(columns = c(g, lower, upper), decimals = 4) |>
  fmt_percent(columns = top1, decimals = 1) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0) |>
  tab_options(table.font.size = px(13), heading.align = "left")
```

### Two-panel concentration figure

```{r}
#| label: fig-lorenz-power
#| fig-width: 12
#| fig-height: 5.8
#| fig-cap: "Capital concentration is near-total in every Brazilian firm-size tier."

# ── Panel A: log-rank concentration curve ───────────────────────────────────
conc_curve <- co |>
  group_by(size) |>
  arrange(desc(capital_stock), .by_group = TRUE) |>
  mutate(
    rank_pct = row_number() / n(),
    cum_top  = cumsum(capital_stock) / sum(capital_stock)
  ) |>
  ungroup() |>
  filter(rank_pct <= 0.5)

p_conc <- ggplot(conc_curve, aes(x = rank_pct, y = cum_top, colour = size)) +
  annotate("segment", x = 0.01, xend = 0.01, y = 0, yend = 1,
           colour = "grey75", linetype = "dashed", linewidth = 0.45) +
  annotate("text", x = 0.01, y = 1.04, label = "Top 1%",
           colour = "grey40", size = 3.3, fontface = "italic") +
  geom_line(linewidth = 1.5, alpha = 0.92) +
  scale_x_log10(
    breaks = c(0.001, 0.01, 0.1, 0.5),
    labels = c("Top 0.1%", "Top 1%", "Top 10%", "Top 50%"),
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  scale_y_continuous(
    labels = percent_format(),
    limits = c(0, 1.08),
    expand = c(0, 0),
    name   = "Cumulative share of tier's capital"
  ) +
  scale_colour_manual(values = pal_size,
                      labels = tier_stats$label,
                      name   = NULL) +
  labs(
    x = "Rank of firms (log scale, highest capital first) →",
    subtitle = "(A)  Concentration curve — log-rank x-axis reveals the\n        tier separation that linear Lorenz cannot show"
  ) +
  theme_cwc() +
  theme(
    legend.position      = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.background    = element_rect(fill = alpha("white", 0.9), colour = NA),
    legend.key.width     = unit(1.5, "lines"),
    legend.text          = element_text(family = "mono", size = 9),
    plot.subtitle        = element_text(colour = "grey35", size = 11,
                                        lineheight = 1.3, margin = margin(b = 10))
  )

# ── Panel B: stacked rank-bucket decomposition ──────────────────────────────
decomp <- co |>
  group_by(size) |>
  arrange(desc(capital_stock), .by_group = TRUE) |>
  mutate(
    rank_pct = row_number() / n(),
    segment  = case_when(
      rank_pct <= 0.01 ~ "Top 1%",
      rank_pct <= 0.10 ~ "Next 9%",
      rank_pct <= 0.50 ~ "Next 40%",
      TRUE             ~ "Bottom 50%"
    ),
    segment  = factor(segment,
                      levels = c("Bottom 50%", "Next 40%", "Next 9%", "Top 1%"))
  ) |>
  group_by(size, segment) |>
  summarise(cap = sum(capital_stock), .groups = "drop") |>
  group_by(size) |>
  mutate(share = cap / sum(cap)) |>
  ungroup()

p_decomp <- ggplot(decomp, aes(x = size, y = share, fill = segment)) +
  geom_col(width = 0.62, colour = "white", linewidth = 1.4) +
  geom_text(
    data = filter(decomp, share >= 0.04),
    aes(label = percent(share, accuracy = 1),
        colour = segment %in% c("Top 1%", "Next 9%")),
    position = position_stack(vjust = 0.5),
    fontface = "bold", size = 4, show.legend = FALSE
  ) +
  scale_fill_manual(
    values = pal_segment,
    breaks = c("Top 1%", "Next 9%", "Next 40%", "Bottom 50%"),
    name   = NULL
  ) +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey25")) +
  scale_y_continuous(labels = percent_format(),
                     expand = c(0, 0),
                     name   = "Share of tier's total capital") +
  labs(
    x = NULL,
    subtitle = "(B)  Each tier's top 1% holds ~96–99% of the tier's\n        capital. The bottom half is a sliver."
  ) +
  theme_cwc() +
  theme(
    legend.position    = "top",
    panel.grid.major.x = element_blank(),
    plot.subtitle      = element_text(colour = "grey35", size = 11,
                                      lineheight = 1.3, margin = margin(b = 10))
  )

# ── Combine ─────────────────────────────────────────────────────────────────
(p_conc | p_decomp) +
  plot_layout(widths = c(1.15, 1)) +
  plot_annotation(
    title    = "Capital concentration is near-total in every Brazilian firm-size tier",
    subtitle = paste0(
      "Gini coefficients exceed 0.98 across Micro, Small, and Medium/Large — ",
      "'micro-enterprise' is not uniquely unequal.\n",
      "The capital trap is a universal power-law structure, not a feature of small firms."
    ),
    caption  = caption_text,
    theme    = theme_cwc()
  ) &
  theme(plot.title = element_text(face = "bold", size = 16, colour = "#1d3557"))
```

### Interpretation

Three findings emerge.

**The Ginis are statistically distinguishable but economically negligible.** All three
tier Ginis exceed 0.98, with bootstrap 95% CIs that are tight and non-overlapping.
Statistically the tiers differ; substantively all three are pinned against the ceiling
of the Gini scale. No reasonable inequality target distinguishes 0.985 from 0.992.

**Top-1% capital share is the dominant fact.** In each tier, the top one percent of
firms holds 96–99% of that tier's capital. The bottom 50% holds less than one percent.
This is not "high inequality" in the colloquial sense; it is a power-law in which
nearly all economic mass sits at the top of the distribution.

**The log-rank view rescues information that linear Lorenz hides.** Panel A shows that
the three tiers are not visually identical when the rank axis is log-scaled — Micro
firms reach 80% cumulative share faster than Medium/Large, indicating that the
*shape* of the tail differs across tiers even when the aggregate Gini is similar.
Linear Lorenz curves cannot show this because the linear x-axis crushes the upper
decile into 10% of the canvas.

The "capital trap" narrative attached to micro-enterprises in earlier framings is
misplaced. The concentration is **structural to the entire Brazilian formal sector**,
not specific to the micro tier. This reframes the analytical agenda: if the inequality
*shape* is universal, the substantive question is no longer *which tier is unequal*
but *who occupies the top of each tier* — and that question cannot be answered by
size classification. It requires governance information.

## Methodological Notes

- **Gini stability at extreme concentration.** When the top firms dominate, Gini
  becomes mechanically close to 1 and economically meaningful changes are compressed
  into the third decimal. Top-1% share is more interpretable in this regime and
  invariant to the lower-tail definition.
- **Bootstrap sample size.** With N > 100,000 per tier, B = 1,000 percentile
  bootstrap resamples produce CIs in the fourth decimal place. We report them as a
  discipline against assuming a single resampling could change the qualitative
  finding, not because they add new information.
- **Excluding zero / missing capital.** Reported zero capital often reflects
  registry-coding artifacts (entities awaiting capital declaration after
  registration) rather than genuinely zero-capital operating firms. Including zeros
  would inflate the bottom-share estimates artifactually and pull the Lorenz curve
  toward exact-L shape.
- **Why not Theil decomposition?** A Theil-T within-vs-between decomposition is a
  natural next step, and earlier drafts of this analysis included one. We dropped
  it here because the universal-power-law finding makes the *between-tier* component
  uninformative: when every tier has Gini > 0.98, decomposing tells us only that
  within-tier inequality dominates, which we already know visually. The decomposition
  is a useful exercise when between-tier differences carry information; here, they
  don't.

## Forward Link

If concentration is universal but the question is *who sits at the top of each tier*,
the analytical next step is to ask whether **governance archetype** — the legal role
of the controlling person — predicts capital independently of size. Part II takes
that question, examining whether governance shifts the *level* of capital, the
*variance* of capital, or both.
