---
title: "Capital Without Category — Part II"
subtitle: "Governance as distributional shape: owner archetype and the heavy-tail signature"
author: "Jerry Miao"
date: "2026-02-25"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-dpi: 180
    fig-align: center
    fig-cap-location: bottom
    df-print: paged
execute:
  warning: false
  message: false
  cache: true
---

## Background

Part I established that capital concentration in Brazilian formal-sector firms is a
universal property: every firm-size tier exhibits Gini > 0.98 and a top-1% capital
share of 96–99%. The shape of the distribution is the same across regulatory
categories. What differs across tiers is *who occupies the top* — and the firm-size
classification cannot answer that question, because it is built on annual revenue and
is silent on the legal architecture of the controlling party.

The CNPJ registry's `owner_qualification` field carries that missing information. It
records the legal role of the controlling person under three broad archetypes:

- **Sole Entrepreneur** — a natural person operating under their own name, fully
  exposed to business risk.
- **Managing Partner / Administrator** — an active operator who runs the business as
  one of one or more named partners with operational authority.
- **Beneficial Owner (passive / investor)** — a controlling equity holder who may sit
  behind a holding-company chain, often without operational involvement.

These archetypes map onto well-established predictions from agency theory. Sole
Entrepreneurs face the highest information asymmetries with external lenders and have
no separation between personal and business liability. Managing Partners signal active
human capital and shared risk-bearing, lowering the information cost of credit.
Beneficial Owners are heterogeneous: at the small end they are dormant shells; at the
large end they are investment vehicles channelling capital from institutional or
family-office structures.

## Research Question

> **Q2.** Conditional on firm size, does governance archetype shape the capital
> distribution? If it does, is the effect a *level shift* (governance moves the
> median), a *variance shift* (governance moves the tail), or both?

The question matters because the answer dictates the targeting strategy. A level
shift means governance acts as a coarse credit-access proxy. A variance shift means
governance is selecting different *populations* of firms — and lumping them by size
will systematically misallocate any size-targeted policy.

## Design Conventions

```{r}
#| label: setup-theme

library(tidyverse)
library(scales)
library(patchwork)
library(gt)

# ── Shared palette ──────────────────────────────────────────────────────────
pal_size <- c(
  "Micro"        = "#c3553a",
  "Small"        = "#d9a441",
  "Medium/Large" = "#2a5a6b"
)

pal_gov <- c(
  "Sole Entrepreneur"                    = "#d9a441",   # amber (solo)
  "Managing Partner /\nAdministrator"    = "#5b8a9a",   # slate blue (active)
  "Beneficial Owner\n(passive/investor)" = "#1d3557"    # deep navy (passive)
)

# ── Shared theme ────────────────────────────────────────────────────────────
theme_cwc <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title        = element_text(face = "bold", size = base_size + 2,
                                       colour = "#1d3557", margin = margin(b = 4)),
      plot.subtitle     = element_text(colour = "grey35", size = base_size - 2,
                                       lineheight = 1.35, margin = margin(b = 18)),
      plot.caption      = element_text(colour = "grey55", size = base_size - 5,
                                       hjust = 0, margin = margin(t = 12)),
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      strip.text        = element_text(face = "bold", size = base_size - 1,
                                       colour = "#1d3557"),
      strip.background  = element_rect(fill = "grey96", colour = NA),
      panel.grid.minor  = element_blank(),
      panel.grid.major  = element_line(colour = "grey92", linewidth = 0.3),
      panel.spacing     = unit(1.4, "lines"),
      legend.position   = "top",
      legend.title      = element_text(face = "bold", size = base_size - 2),
      legend.text       = element_text(size = base_size - 2),
      axis.title        = element_text(colour = "grey30", size = base_size - 2),
      axis.text         = element_text(colour = "grey40", size = base_size - 3)
    )
}

caption_text <- paste(
  "Source: Brazilian CNPJ Registry · TidyTuesday 2026 Wk 4 (Extended)",
  "· Boyd, He & Miao · 2026-02-25"
)
```

## Data Preparation

```{r}
#| label: data-prep

companies <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv",
  show_col_types = FALSE
)

co <- companies |>
  filter(
    !is.na(company_size),
    !is.na(capital_stock),
    !is.na(owner_qualification),
    capital_stock > 0
  ) |>
  mutate(
    size = case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE                               ~ "Medium/Large"
    ),
    size = factor(size, levels = c("Micro", "Small", "Medium/Large")),

    governance = case_when(
      str_detect(owner_qualification,
                 regex("beneficial owner", ignore_case = TRUE))     ~ "Beneficial Owner\n(passive/investor)",
      str_detect(owner_qualification,
                 regex("entrepreneur|business owner", ignore_case = TRUE)) ~ "Sole Entrepreneur",
      str_detect(owner_qualification,
                 regex("managing partner|administrator", ignore_case = TRUE)) ~ "Managing Partner /\nAdministrator",
      TRUE ~ "Other"
    ),
    governance = factor(governance, levels = c(
      "Sole Entrepreneur",
      "Managing Partner /\nAdministrator",
      "Beneficial Owner\n(passive/investor)",
      "Other"
    )),

    log_capital = log10(capital_stock)
  )

cat("Observations:", comma(nrow(co)), "\n")
cat("Governance archetypes:\n"); print(count(co, governance, sort = TRUE))
```

## Descriptive Snapshot: Governance × Size Cells

The table below carries the data plotted in the two-panel figure that follows.
Two columns matter most: the **median**, which gives the level summary that
Panel A tracks across tiers, and the **mean / median** ratio, which gives the
tail-heaviness summary that Panel B tracks. Ratios close to 1 signal a roughly
symmetric distribution; ratios in the double or triple digits signal a heavy
right tail. As we will see, both moments tell tier-dependent stories rather
than uniform archetype-level ones.

```{r}
#| label: tbl-governance-cells

gov_cells <- co |>
  filter(governance != "Other") |>
  group_by(size, governance) |>
  summarise(
    n         = n(),
    median    = median(capital_stock),
    mean      = mean(capital_stock),
    p90       = quantile(capital_stock, 0.90),
    mean_med  = mean / median,
    .groups   = "drop"
  )

gov_cells |>
  mutate(governance = str_replace(governance, "\n", " ")) |>
  gt(groupname_col = "size") |>
  tab_header(
    title    = md("**Capital stock by size × governance**"),
    subtitle = "Mean / median ratio diagnoses tail heaviness within each cell"
  ) |>
  cols_label(
    governance = "Governance archetype",
    n = "N", median = "Median", mean = "Mean",
    p90 = "P90", mean_med = "Mean / Median"
  ) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = c(median, mean, p90),
             suffixing = TRUE, decimals = 1) |>
  fmt_number(columns = mean_med, decimals = 1) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) |>
  tab_options(
    table.font.size            = px(13),
    heading.align              = "left",
    row_group.background.color = "#f4f4f4"
  ) |>
  tab_source_note(md("*Source: Brazilian CNPJ Registry*"))
```

## Analysis: Governance Shapes Distributions, Not Just Levels

### Methodological motivation

A boxplot or a bar chart of medians would already show that governance archetypes
differ on average — but it would miss the analytically critical question: do they
differ in *level* (a translation of the entire distribution) or in *shape*
(different variances and tail heaviness)? Two complementary panels surface both
moments at once:

- **Panel A — median trajectories with range.** For each governance archetype,
  plot median capital across size tiers, with the middle 80% of the distribution
  (P10–P90) shown as thin whiskers and the IQR (P25–P75) as a thicker
  translucent bar. Connecting medians across tiers within an archetype reveals
  whether that archetype's *level* scales with size or stays flat.
- **Panel B — mean ÷ median ratio.** The gap between mean and median is a robust
  tail-heaviness diagnostic: when the right tail is heavy, the mean is pulled
  far above the median and the ratio diverges from 1. Plotting this ratio on a
  log scale by size tier surfaces tail behaviour that the median trajectory in
  Panel A cannot show directly.

Why not density ridges? An earlier draft used `geom_density_ridges` to show full
distributional shape side-by-side. Ridges work well when distributions are
roughly comparable in scale, but capital stock is so heavy-tailed that the ridge
for each cell collapses into a thin spike at the lower end with a faint
right-trailing arm — visually, all nine cells look nearly identical even when
their statistical moments differ by orders of magnitude. The trajectory + ratio
design isolates the two moments that actually carry the story.

For inferential support we use the **Kruskal-Wallis test** within each size tier.
This is a non-parametric one-way analysis of variance on ranks: appropriate here
because capital-stock distributions are extremely log-normal at best and have
heavy right tails that violate ANOVA's normality assumption. Kruskal-Wallis
tests the null hypothesis that all governance subgroups within a size tier are
drawn from the same distribution. Rejecting the null does not by itself say
*what* differs — that interpretation comes from the visual.

### Kruskal-Wallis tests within each size tier

```{r}
#| label: kw-test

kw_results <- co |>
  filter(governance != "Other") |>
  group_by(size) |>
  summarise(
    statistic = kruskal.test(capital_stock ~ governance)$statistic,
    df        = kruskal.test(capital_stock ~ governance)$parameter,
    p_value   = kruskal.test(capital_stock ~ governance)$p.value,
    n         = n(),
    .groups   = "drop"
  )

kw_results |>
  gt() |>
  tab_header(
    title    = md("**Kruskal-Wallis tests of governance × capital within each tier**"),
    subtitle = "H₀: all three governance subgroups draw from the same capital distribution"
  ) |>
  cols_label(
    size = "Tier", statistic = "H statistic",
    df = "df", p_value = "p-value", n = "N"
  ) |>
  fmt_number(columns = c(statistic), decimals = 1) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0) |>
  fmt_scientific(columns = p_value, decimals = 2) |>
  tab_options(table.font.size = px(13), heading.align = "left")
```

The null is rejected at conventional levels in every tier. Governance and capital
are not independent within any size category. The visual now supplies the *what*.

### Two-panel figure: median trajectories + tail-heaviness ratio

```{r}
#| label: fig-governance-shape
#| fig-width: 12
#| fig-height: 5.8
#| fig-cap: "Governance shapes the capital distribution differently across archetypes."

# ── Per-cell summary statistics ─────────────────────────────────────────────
gov_summary <- co |>
  filter(governance != "Other") |>
  group_by(size, governance) |>
  summarise(
    p10    = quantile(capital_stock, 0.10),
    p25    = quantile(capital_stock, 0.25),
    median = median(capital_stock),
    mean   = mean(capital_stock),
    p75    = quantile(capital_stock, 0.75),
    p90    = quantile(capital_stock, 0.90),
    n      = n(),
    .groups = "drop"
  ) |>
  mutate(
    mean_med  = mean / median,
    gov_short = case_when(
      str_detect(governance, "Beneficial") ~ "Beneficial Owner",
      str_detect(governance, "Managing")   ~ "Managing Partner",
      TRUE                                  ~ "Sole Entrepreneur"
    ),
    gov_short = factor(gov_short, levels = c(
      "Sole Entrepreneur", "Managing Partner", "Beneficial Owner"
    ))
  )

pal_gov_short <- c(
  "Sole Entrepreneur" = "#d9a441",
  "Managing Partner"  = "#5b8a9a",
  "Beneficial Owner"  = "#1d3557"
)

dodge_w <- 0.45

# ── Panel A: median trajectories with IQR + P10–P90 whiskers ────────────────
p_level <- ggplot(
  gov_summary,
  aes(x = size, y = median, colour = gov_short, group = gov_short)
) +
  geom_linerange(aes(ymin = p10, ymax = p90),
                 linewidth = 0.7, alpha = 0.50,
                 position = position_dodge(width = dodge_w)) +
  geom_linerange(aes(ymin = p25, ymax = p75),
                 linewidth = 4.2, alpha = 0.30,
                 position = position_dodge(width = dodge_w)) +
  geom_line(linewidth = 1.3, alpha = 0.85,
            position = position_dodge(width = dodge_w)) +
  geom_point(size = 5, shape = 21, fill = "white", stroke = 2,
             position = position_dodge(width = dodge_w)) +
  scale_y_log10(
    labels = label_number(scale_cut = cut_short_scale(), accuracy = 1),
    breaks = c(1e5, 3e5, 1e6, 3e6, 1e7),
    name   = "Capital stock (BRL, log scale)"
  ) +
  scale_colour_manual(
    values = pal_gov_short, name = NULL,
    guide  = guide_legend(override.aes = list(linewidth = 1.5, size = 4))
  ) +
  labs(
    x = NULL,
    subtitle = "(A)  Median capital with IQR (thick) and P10–P90 (thin)\n        Only Managing Partners scale upward across tiers"
  ) +
  theme_cwc() +
  theme(
    legend.position    = "top",
    panel.grid.major.x = element_blank(),
    plot.subtitle      = element_text(colour = "grey35", size = 11,
                                      lineheight = 1.3, margin = margin(b = 10))
  )

# ── Panel B: mean ÷ median ratio (tail heaviness) ──────────────────────────
p_tail <- ggplot(
  gov_summary,
  aes(x = size, y = mean_med, colour = gov_short, group = gov_short)
) +
  geom_hline(yintercept = 1, linewidth = 0.4,
             colour = "grey70", linetype = "dashed") +
  geom_line(linewidth = 1.3, alpha = 0.85) +
  geom_point(size = 5, shape = 21, fill = "white", stroke = 2) +
  # Label only the rightmost (Medium/Large) endpoint of each line
  # — avoids label collision at Micro/Small where Sole Entrepreneur and
  # Managing Partner ratios are similar
  geom_text(
    data = filter(gov_summary, size == "Medium/Large"),
    aes(label = paste0(round(mean_med), "×")),
    hjust = -0.4, fontface = "bold", size = 4.2, show.legend = FALSE
  ) +
  scale_y_log10(
    breaks = c(1, 3, 10, 30, 100),
    labels = function(x) paste0(x, "×"),
    expand = expansion(mult = c(0.05, 0.22))
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.10, 0.22))) +
  scale_colour_manual(values = pal_gov_short, guide = "none") +
  coord_cartesian(clip = "off") +
  labs(
    x = NULL,
    y = "Mean ÷ median  (tail heaviness)",
    subtitle = "(B)  Mean-to-median ratio\n        Tails are heavy in every cell, but the carrier shifts across tiers"
  ) +
  theme_cwc() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.subtitle      = element_text(colour = "grey35", size = 11,
                                      lineheight = 1.3, margin = margin(b = 10)),
    plot.margin        = margin(10, 25, 10, 10)
  )

# ── Combine ────────────────────────────────────────────────────────────────
(p_level | p_tail) +
  plot_layout(widths = c(1.4, 1), guides = "collect") +
  plot_annotation(
    title    = "Managing Partners scale the median; capital tails migrate across archetypes",
    subtitle = paste0(
      "Capital distribution by owner archetype within each firm-size tier. ",
      "Panel A: median trajectory with middle 80% range.\n",
      "Panel B: mean ÷ median ratio captures tail heaviness — BO carries the tail at Micro; MP carries it at Medium/Large."
    ),
    caption  = caption_text,
    theme    = theme_cwc()
  ) &
  theme(
    plot.title      = element_text(face = "bold", size = 16, colour = "#1d3557"),
    legend.position = "top"
  )
```

### Interpretation

Panel A's three trajectories give the level story:

- **Managing Partners** are the only archetype whose median materially scales
  with size: **300k → 360k → 1M** (a 3.3× increase from Micro to Medium/Large).
  Active operators accumulate capital as firms grow — capital and operational
  scale move together.
- **Sole Entrepreneurs** show a flat median across Micro and Small (~300k),
  with a modest uptick at Medium/Large (~394k). Solo proprietors face
  information asymmetries that cap external capital access; revenue scale does
  not translate into capital in the same way.
- **Beneficial Owners** stay flat at 300k throughout — for this archetype, the
  median is uninformative about firm size.

Panel B's mean ÷ median ratios complicate the simple "Beneficial Owners carry
the tail" story we initially expected. Capital distributions are heavy-tailed
in nearly every cell (most ratios fall between 30× and 500×), but **which
archetype carries the heaviest tail changes with firm size**:

- **At Micro**, Beneficial Owners share the heaviest tail with Managing
  Partners (~80× each), while Sole Entrepreneurs are the lightest (~30×). This
  is the "capital-rich micro" pattern from the project framing: most BO micros
  are tiny, but a holding-company minority hold enormous capital.
- **At Small**, the pattern reorganizes. Managing Partners' tail grows sharply
  (~300×), Beneficial Owners' tail moderates (~50×), and Sole Entrepreneurs'
  tail collapses to near-symmetry (~2×). The Small tier appears to discipline
  capital distributions — possibly because Brazil's revenue-band classifications
  create selection pressures at the boundaries.
- **At Medium/Large**, Managing Partners reach an extraordinary **526×** ratio,
  Sole Entrepreneurs a **127×** ratio, and Beneficial Owners drop to **3×** —
  the *lightest* tail in the entire panel. The largest active-management firms
  include genuine industry titans whose capital dwarfs the typical MP firm.
  Beneficial Owner firms classified as Medium/Large are, by contrast, a much
  more uniform population — likely dominated by institutional investment
  vehicles rather than the bimodal small-shell + mega-vehicle distribution
  found at Micro.

The structural finding is richer than a uniform "BO heavy-tail" story:

- **Capital-tail composition migrates across archetypes as firms scale.** At
  the small end, the tail belongs to passive ownership (holding-company shells
  using BO governance). At the large end, the tail belongs to active
  management (true industry champions). The two heavy tails point to two
  structurally different phenomena that should not be conflated.
- **The "capital-rich micro" finding is specifically a Micro × Beneficial
  Owner phenomenon.** It does not generalize to other tiers — and the
  heavy-tailed inequality at Medium/Large is a different question (extreme
  inequality among genuine large firms) that would need its own analytical
  frame.
- **Sole Entrepreneurs are not flat in the strong sense.** Their median is
  roughly flat, but their tail is volatile — collapsing at Small and resurging
  at Medium/Large. The V-shape may reflect regulatory-boundary effects of
  Brazil's revenue classifications, where firms strategically position
  themselves at specific capital levels to access (or avoid) tier-specific
  tax treatments.

The governance premium is not uniform along any single axis. At each tier it
has a different shape, and any size-targeted policy that treats these
archetypes as equivalent within a tier will systematically misallocate. The
next step — Part III — zooms in on the Micro tier specifically, where the
BO-tail phenomenon is sharpest, to identify the legal-form channel through
which capital-rich micros achieve their unusual endowment.

## Methodological Notes

- **Why Kruskal-Wallis rather than ANOVA.** Capital-stock distributions are
  strongly right-skewed even after log transformation; ANOVA's normality
  assumption is violated. Kruskal-Wallis operates on ranks and is robust to
  this.
- **Why a two-panel design rather than ridges.** Ridge plots show the full
  distribution shape, which is appealing in principle. In practice,
  capital-stock distributions are so heavy-tailed that the ridge for each cell
  collapses into a near-spike at the lower end, with the upper tail rendered
  as a faint trailing arm at any reasonable bandwidth. All nine size ×
  governance cells then look visually similar even when their statistical
  moments differ by orders of magnitude. The trajectory + tail-ratio
  decomposition isolates the two moments (level, tail) that actually carry the
  analytical story.
- **Why mean-to-median ratio as the tail summary.** The ratio is scale-invariant
  (insensitive to currency unit), bounded below by 1 for symmetric
  distributions, and easy to read as "the mean is N times the median".
  Alternatives like skewness or kurtosis are mathematically equivalent in
  spirit but harder for non-statistical readers to interpret directly.
- **Sensitivity of the ratio to extreme values.** The mean is by construction
  pulled by upper-tail outliers, which makes the ratio a *combined* signal of
  tail heaviness *and* outlier presence. At Medium/Large, where each
  governance cell contains thousands rather than hundreds of thousands of
  firms, a small number of mega-corporations can move the ratio dramatically.
  The 526× value for Managing Partners at Medium/Large reflects a real
  population-level fact (a handful of firms with capital in the hundreds of
  millions of BRL), but its precise magnitude could shift across years or
  registry vintages. The qualitative ranking — MP heaviest, BO lightest at
  Medium/Large — is robust; the specific multiples should be read as
  order-of-magnitude indicators rather than precise estimates.
- **Excluding "Other" governance.** A residual category with heterogeneous
  coding (state-owned representatives, foreign legal entities, etc.) is
  excluded from the figure to keep archetype-level interpretation clean. It is
  approximately 6% of observations.
- **Why we dropped quantile regression here.** An earlier draft estimated
  τ ∈ {0.10, 0.25, 0.50, 0.75, 0.90} regressions. The two-panel design
  communicates the same finding (premium widens at the upper tail) more
  directly and without the rank-conditioned interpretation that quantile
  regression imposes. The distributional shape *is* the result.

## Forward Link

The most policy-relevant finding from this analysis is the **Micro × Beneficial
Owner** cell: at the smallest tier, Beneficial Owners hold the heaviest tail,
which is the statistical fingerprint of the "capital-rich micro" phenomenon
that motivates this project. The development-policy question that follows is
*which legal-form channel* turns that ownership pattern into a registered
micro-enterprise with above-tier-typical capital. Part III ranks the
legal-form × governance cells inside the micro tier and identifies the
partnership-with-passive-equity structure as the dominant escape route — a
result that overturns the popular "EIRELI-as-escape-route" framing. The
Medium/Large tail behaviour (where Managing Partners reach a 526× mean/median
ratio) is a different empirical phenomenon — extreme inequality among genuine
large operating firms — and is not pursued further here, since it lies outside
the small-business policy remit of this trilogy.
