---
title: "Member 3 Jerry Miao"
---

## Data and Setup

### Packages

```{r}
#| label: setup

library(tidyverse)
library(scales)
library(ggridges)
library(patchwork)
library(boot)          # bootstrap Gini confidence intervals
library(quantreg)      # quantile regression
library(broom)         # tidy model outputs
library(gt)            # publication-quality tables
library(modelsummary)  # regression summary tables
library(ggtext)        # markdown / HTML in ggplot labels
library(glue)          # string interpolation

# ── Shared palette ─────────────────────────────────────────────────────────────
pal_size <- c("Micro" = "#e76f51", "Small" = "#f4a261", "Medium/Large" = "#264653")
pal_gov  <- c(
  "Sole Entrepreneur"                    = "#e9c46a",
  "Managing Partner /\nAdministrator"    = "#4a90d9",
  "Beneficial Owner\n(passive/investor)" = "#264653"
)

caption_base <- "Source: Brazilian CNPJ Registry ·He, Miao & Boyd"
```

### Data Ingestion and Cleaning

```{r}
#| label: data-load

companies <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv",
  show_col_types = FALSE
)

# ── Core recoding ───────────────────────────────────────────────────────────────
co <- companies |>
  # Drop records with missing or non-positive capital (not usable for log-scale analysis)
  filter(
    !is.na(company_size),
    !is.na(capital_stock),
    capital_stock > 0,
    !is.na(owner_qualification)
  ) |>
  mutate(
    # Size tier (ordered factor)
    size = case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE                               ~ "Medium/Large"
    ),
    size = factor(size, levels = c("Micro", "Small", "Medium/Large")),

    # Governance archetype: collapse owner_qualification into 4 classes
    governance = case_when(
      str_detect(owner_qualification,
                 regex("beneficial owner", ignore_case = TRUE)) ~
        "Beneficial Owner\n(passive/investor)",
      str_detect(owner_qualification,
                 regex("entrepreneur|business owner", ignore_case = TRUE)) ~
        "Sole Entrepreneur",
      str_detect(owner_qualification,
                 regex("managing partner|administrator", ignore_case = TRUE)) ~
        "Managing Partner /\nAdministrator",
      TRUE ~ "Other"
    ),
    governance = factor(governance, levels = c(
      "Sole Entrepreneur",
      "Managing Partner /\nAdministrator",
      "Beneficial Owner\n(passive/investor)",
      "Other"
    )),

    # Log10 capital for distributional analysis
    log_capital = log10(capital_stock),

    # Legal-form shorthand for micro analysis
    legal_short = case_when(
      str_detect(legal_nature, "LLC")                 ~ "LLC",
      str_detect(legal_nature, "Sole Proprietorship") ~ "Sole Prop.",
      str_detect(legal_nature, "Individual Limited")  ~ "EIRELI",
      str_detect(legal_nature, "Simple")              ~ "Simple Part.",
      str_detect(legal_nature, "Silent")              ~ "Silent Part.",
      str_detect(legal_nature, "Corporation")         ~ "Corporation",
      TRUE                                            ~ "Other"
    ),

    # Capital-rich flag: top quartile *within* micro-enterprises
    capital_q75_micro     = quantile(capital_stock[size == "Micro"], 0.75, na.rm = TRUE),
    is_capital_rich_micro = (size == "Micro") & (capital_stock >= capital_q75_micro)
  )

cat("Final observations:", scales::comma(nrow(co)), "\n")
```
---

## Analysis III — The Capital-Rich Micro: Who Escapes the Trap? {#sec-escape}

### Escape Rates by Legal Form x Governance

```{r}
#| label: escape-rates
#| fig-width: 11
#| fig-height: 5.5

micro_profile <- co |>
  filter(size == "Micro", governance != "Other", legal_short != "Other")

escape_rate <- micro_profile |>
  group_by(legal_short, governance) |>
  summarise(
    n_total    = n(),
    n_rich     = sum(is_capital_rich_micro),
    escape_pct = n_rich / n_total,
    # Wilson score confidence interval for a proportion
    ci_lo = {
      p <- escape_pct; n <- n_total; z <- 1.96
      (p + z^2 / (2 * n) - z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))) /
        (1 + z^2 / n)
    },
    ci_hi = {
      p <- escape_pct; n <- n_total; z <- 1.96
      (p + z^2 / (2 * n) + z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))) /
        (1 + z^2 / n)
    },
    .groups = "drop"
  ) |>
  filter(n_total >= 30)

ggplot(
  escape_rate,
  aes(
    x    = fct_reorder(legal_short, escape_pct, .fun = mean),
    y    = escape_pct,
    fill = governance
  )
) +
  geom_col(position = position_dodge(width = 0.75),
           width = 0.65, alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lo, ymax = ci_hi),
    position = position_dodge(width = 0.75),
    width = 0.22, colour = "grey35", linewidth = 0.55
  ) +
  geom_text(
    aes(label = percent(escape_pct, accuracy = 0.1), y = ci_hi),
    position = position_dodge(width = 0.75),
    hjust = -0.15, size = 2.8, colour = "grey25"
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    labels = percent_format(),
    limits = c(0, 0.60),
    expand = expansion(mult = c(0, 0.05)),
    name   = "Share above 75th capital percentile (Wilson 95% CI)"
  ) +
  scale_fill_manual(values = pal_gov, name = "Governance archetype") +
  labs(
    title    = "Beneficial Owner-controlled LLCs are most likely to escape the capital trap",
    subtitle = glue(
      "Among micro-enterprises only (N = {comma(nrow(micro_profile))}).\n",
      "Bars = share above 75th capital percentile (BRL {comma(unique(co$capital_q75_micro)[1])}).",
      " Error bars = Wilson 95% CI. Cells with N < 30 excluded."
    ),
    x       = NULL,
    caption = caption_base
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", size = 14, colour = "#264653"),
    plot.subtitle      = element_text(colour = "grey40", size = 10.5,
                                      lineheight = 1.3, margin = margin(b = 14)),
    plot.caption       = element_text(colour = "grey50", size = 8, hjust = 0),
    legend.position    = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

### Logistic Regression: Odds Ratios for Capital-Rich Status

The bar chart above shows unconditional escape rates. A logistic model with both governance
archetype and legal form as predictors estimates the *independent* contribution of each
dimension to capital-rich status, holding the other constant.

```{r}
#| label: logistic-regression
#| fig-width: 10
#| fig-height: 5.5

logit_fit <- glm(
  is_capital_rich_micro ~ governance + legal_short,
  data = micro_profile |>
    mutate(
      governance  = relevel(factor(governance),  ref = "Sole Entrepreneur"),
      legal_short = relevel(factor(legal_short), ref = "Sole Prop.")
    ),
  family = binomial(link = "logit")
)

or_df <- tidy(logit_fit, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    category   = if_else(str_detect(term, "^governance"), "Governance", "Legal Form"),
    term_clean = str_remove(term, "^governance|^legal_short"),
    term_clean = str_replace_all(term_clean, "\\.", " "),
    sig_label  = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    or_label = glue("{round(estimate, 2)}{sig_label}")
  )

ggplot(or_df,
       aes(x = estimate,
           y = fct_reorder(term_clean, estimate),
           colour = category)) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "grey60", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.3, linewidth = 0.8) +
  geom_point(size = 3.5) +
  geom_text(aes(label = or_label, x = conf.high),
            hjust = -0.2, size = 3.2, colour = "grey25") +
  facet_wrap(~category, scales = "free_y", ncol = 2) +
  scale_colour_manual(
    values = c("Governance" = "#264653", "Legal Form" = "#4a90d9"),
    guide  = "none"
  ) +
  scale_x_log10(
    name   = "Odds Ratio (log scale)  —  ref: Sole Entrepreneur x Sole Prop.",
    labels = \(x) round(x, 2)
  ) +
  labs(
    title    = "Beneficial Ownership and LLC form are the strongest predictors of capital-rich micro status",
    subtitle = paste0(
      "Logistic regression odds ratios with 95% profile likelihood CI.\n",
      "Outcome: micro-enterprise above 75th capital percentile.\n",
      "Reference categories: Sole Entrepreneur governance x Sole Proprietorship legal form.\n",
      "Significance stars: *** p<0.001  ** p<0.01  * p<0.05"
    ),
    y       = NULL,
    caption = caption_base
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 13, colour = "#264653"),
    plot.subtitle    = element_text(colour = "grey40", size = 10.2,
                                    lineheight = 1.3, margin = margin(b = 14)),
    plot.caption     = element_text(colour = "grey50", size = 8, hjust = 0),
    strip.text       = element_text(face = "bold", size = 11, colour = "#264653"),
    strip.background = element_rect(fill = "grey95", colour = NA),
    panel.grid.minor = element_blank()
  )
```

---

## Regression Table: Nested OLS Models of Log Capital

The logistic model above is restricted to micro-enterprises. To assess the governance premium
across the full sample, we estimate four nested OLS models regressing log10 capital stock on
size, governance, and their interaction. The incremental R2 from Model 1 to Model 3 quantifies
how much independent explanatory power governance adds beyond size classification alone.

```{r}
#| label: ols-table

# Recode governance to short labels before fitting so coefficient names are clean
reg_data <- co |>
  filter(governance != "Other") |>
  mutate(
    gov = fct_recode(governance,
      "Managing Partner" = "Managing Partner /\nAdministrator",
      "Beneficial Owner" = "Beneficial Owner\n(passive/investor)",
      "Sole Entrepreneur" = "Sole Entrepreneur"
    ),
    gov = relevel(gov, ref = "Sole Entrepreneur")
  )

m1 <- lm(log_capital ~ size,      data = reg_data)
m2 <- lm(log_capital ~ gov,       data = reg_data)
m3 <- lm(log_capital ~ size + gov, data = reg_data)

modelsummary(
  list("Size only" = m1, "Governance only" = m2, "Size + Governance" = m3),
  stars      = TRUE,
  coef_rename = c(
    "(Intercept)"            = "Intercept",
    "sizeSmall"              = "Size: Small",
    "sizeMedium/Large"       = "Size: Medium / Large",
    "govManaging Partner"    = "Gov: Managing Partner",
    "govBeneficial Owner"    = "Gov: Beneficial Owner"
  ),
  gof_map = c("nobs", "r.squared", "adj.r.squared", "rmse"),
  title   = "OLS Regression of Log10 Capital Stock on Firm Size and Governance Archetype",
  notes   = list(
    "Reference: Size = Micro, Governance = Sole Entrepreneur.",
    "DV: log10(capital_stock). SEs in parentheses. * p<0.05  ** p<0.01  *** p<0.001"
  )
)
```

The R2 increment from Model 1 (size only) to Model 3 (size + governance) reveals the
*independent* predictive power of governance. Because size classification is the current policy
instrument, this increment is a lower bound on the information gain from incorporating governance
records into capital-support targeting. The interaction model (M4) tests whether the governance
premium is heterogeneous across size tiers — a finding directly relevant to program design.

---
