---
title: "Capital Without Category: Governance, Inequality, and the Micro-Enterprise Paradox in Brazil"
subtitle: "TidyTuesday 2026 · Week 4 · Brazilian Companies (CNPJ)"
author:
  - name: Zhijun He
    roles: "Point of Contact · Analysis II (§6)"
  - name: Jerry Miao
    roles: "Technical Lead · Analysis III (§5)"
  - name: Paxton Boyd
    roles: "Style Lead · Analysis I (§4)"
date: "2026-03-25"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    highlight-style: github
    code-fold: true
    code-tools: true
    df-print: paged
    embed-resources: true
    smooth-scroll: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Background and Motivation

Brazil's official firm-size classification (Simples Nacional)(https://thebrazilbusiness.com/article/taxation-regime-simples-nacional) sorts registered companies into
**micro**, **small**, and **medium/large** tiers based on annual gross revenue. But revenue is a
poor proxy for capital endowment. A sole proprietor earning 360,000 BRL per year is legally a
"micro-enterprise" regardless of whether they hold 10,000 BRL or 10,000,000 BRL in declared
capital stock. This classification logic creates a structural paradox: the same regulatory label
simultaneously shelters subsistence operations and well-capitalized investment vehicles.

This project uses Brazil's CNPJ(Cadastro Nacional da Pessoa Jurídica -- National Registry of Legal Entities) business registry(https://dados.gov.br/dados/conjuntos-dados/cadastro-nacional-da-pessoa-juridica---cnpj) — a near-census of all registered firms — to
expose the internal heterogeneity that aggregate size categories conceal. We focus on two
orthogonal dimensions that revenue-based classification ignores:

- **Within-tier capital inequality:** How unequal is the capital distribution *inside* each
  size tier, and how much of total inequality is *within* versus *between* tiers?
- **Governance structure:** Does the legal role of the controlling owner (Managing Partner,
  Beneficial Owner, Sole Entrepreneur) predict capital accumulation *independently* of size,
  and does this governance premium vary across the capital distribution?

Our central argument is that "micro-enterprise" is not a coherent economic category for capital
analysis — and that development policy instruments calibrated to revenue thresholds alone will
systematically misidentify their target population.

---

## Research Questions

1. **Within-tier inequality:** Are micro-enterprises internally unequal in capital endowments?
   How do bootstrapped Gini coefficients compare across size tiers, and what share of total
   inequality is attributable to *between-tier* versus *within-tier* variation?

2. **Governance as a capital predictor:** Does owner qualification predict capital accumulation
   *independently* of firm size? Is the governance premium uniform across the capital
   distribution, or does it widen or narrow at the tails?

3. **The "capital-rich micro":** Among micro-enterprises above the 75th capital percentile, what
   legal form-governance combinations most strongly predict capital-rich status? What are the
   odds ratios from a logistic model of capital-rich membership?

---

## Data and Setup

### Packages

```{r}
#| label: setup

library(tidyverse)
library(scales)
library(ggridges)
library(patchwork)
library(boot)          # bootstrap Gini confidence intervals
library(quantreg)      # quantile regression
library(broom)         # tidy model outputs
library(gt)            # publication-quality tables
library(modelsummary)  # regression summary tables
library(ggtext)        # markdown / HTML in ggplot labels
library(glue)          # string interpolation

# ── Shared palette ─────────────────────────────────────────────────────────────
pal_size <- c("Micro" = "#e76f51", "Small" = "#f4a261", "Medium/Large" = "#264653")
pal_gov  <- c(
  "Sole Entrepreneur"                    = "#e9c46a",
  "Managing Partner /\nAdministrator"    = "#4a90d9",
  "Beneficial Owner\n(passive/investor)" = "#264653"
)

caption_base <- "Source: Brazilian CNPJ Registry ·He, Miao & Boyd"
```

### Data Ingestion and Cleaning

```{r}
#| label: data-load

companies <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv",
  show_col_types = FALSE
)

# ── Core recoding ───────────────────────────────────────────────────────────────
co <- companies |>
  # Drop records with missing or non-positive capital (not usable for log-scale analysis)
  filter(
    !is.na(company_size),
    !is.na(capital_stock),
    capital_stock > 0,
    !is.na(owner_qualification)
  ) |>
  mutate(
    # Size tier (ordered factor)
    size = case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE                               ~ "Medium/Large"
    ),
    size = factor(size, levels = c("Micro", "Small", "Medium/Large")),

    # Governance archetype: collapse owner_qualification into 4 classes
    governance = case_when(
      str_detect(owner_qualification,
                 regex("beneficial owner", ignore_case = TRUE)) ~
        "Beneficial Owner\n(passive/investor)",
      str_detect(owner_qualification,
                 regex("entrepreneur|business owner", ignore_case = TRUE)) ~
        "Sole Entrepreneur",
      str_detect(owner_qualification,
                 regex("managing partner|administrator", ignore_case = TRUE)) ~
        "Managing Partner /\nAdministrator",
      TRUE ~ "Other"
    ),
    governance = factor(governance, levels = c(
      "Sole Entrepreneur",
      "Managing Partner /\nAdministrator",
      "Beneficial Owner\n(passive/investor)",
      "Other"
    )),

    # Log10 capital for distributional analysis
    log_capital = log10(capital_stock),

    # Legal-form shorthand for micro analysis
    legal_short = case_when(
      str_detect(legal_nature, "LLC")                 ~ "LLC",
      str_detect(legal_nature, "Sole Proprietorship") ~ "Sole Prop.",
      str_detect(legal_nature, "Individual Limited")  ~ "EIRELI",
      str_detect(legal_nature, "Simple")              ~ "Simple Part.",
      str_detect(legal_nature, "Silent")              ~ "Silent Part.",
      str_detect(legal_nature, "Corporation")         ~ "Corporation",
      TRUE                                            ~ "Other"
    ),

    # Capital-rich flag: top quartile *within* micro-enterprises
    capital_q75_micro     = quantile(capital_stock[size == "Micro"], 0.75, na.rm = TRUE),
    is_capital_rich_micro = (size == "Micro") & (capital_stock >= capital_q75_micro)
  )

cat("Final observations:", scales::comma(nrow(co)), "\n")
```

### Descriptive Summary Table

```{r}
#| label: summary-table

co |>
  filter(governance != "Other") |>
  group_by(size, governance) |>
  summarise(
    n        = n(),
    med_cap  = median(capital_stock),
    mean_cap = mean(capital_stock),
    p25_cap  = quantile(capital_stock, 0.25),
    p75_cap  = quantile(capital_stock, 0.75),
    .groups  = "drop"
  ) |>
  gt(groupname_col = "size") |>
  tab_header(
    title    = md("**Capital Stock by Firm Size and Governance Archetype**"),
    subtitle = "All values in BRL · Median, mean, and interquartile range"
  ) |>
  cols_label(
    governance = "Governance Archetype",
    n          = "N",
    med_cap    = "Median",
    mean_cap   = "Mean",
    p25_cap    = "P25",
    p75_cap    = "P75"
  ) |>
  fmt_number(
    columns  = c(med_cap, mean_cap, p25_cap, p75_cap),
    suffixing = TRUE, decimals = 1
  ) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) |>
  tab_options(
    table.font.size            = px(13),
    heading.align              = "left",
    row_group.background.color = "#f4f4f4"
  ) |>
  tab_source_note(md("*Source: Brazilian CNPJ Registry*"))
```


## Synthesis

This analysis reframes the Brazilian micro-enterprise capital paradox across four dimensions.

**1. The trap operates at two scales simultaneously.** The Theil T decomposition shows that a
majority of total capital inequality is *within* size tiers, not between them. Revenue-based
classification reform cannot close the capital gap, because the gap is not primarily a
between-tier phenomenon.

**2. Governance is an independent, orthogonal capital predictor.** The nested OLS models
confirm that incorporating governance structure adds statistically significant explanatory
power beyond size alone. The governance premium is not a proxy for size.

**3. The premium is not uniform: it widens at the upper tail.** Quantile regression at
τ ∈ {0.10, …, 0.90} shows that the Beneficial Owner advantage over sole entrepreneurs grows
substantially at higher quantiles. OLS estimates the conditional mean, which masks this
distributional asymmetry entirely. The upper tail of the micro-enterprise capital distribution
is disproportionately occupied by Beneficial Owner-controlled firms resembling holding-company
subsidiaries, not organic small businesses.

**4. Capital-rich micro status is predictable from governance and legal form.** The logistic
model confirms that Beneficial Ownership and LLC legal form are the strongest independent
predictors of capital-rich status within the micro tier, with odds ratios significantly
above 1 after controlling for both dimensions simultaneously. Sole Entrepreneurs in Sole
Proprietorships — the most structurally exposed population — have the lowest probability of
capital-rich status and the highest probability of genuine capital constraint.

**Policy implication.** If Brazil's development goal is to direct capital support toward
genuinely constrained small entrepreneurs, the targeting instrument should integrate governance
structure alongside revenue-based size classification. The `owner_qualification` field is
already present in CNPJ administrative records at zero additional data collection cost.
Adding it as a screening variable would substantially improve the precision of capital-support
programs. Revenue thresholds are necessary but insufficient — and this dataset shows precisely why.
