---
title: "Capital Without Category: Governance, Inequality, and the Micro-Enterprise Paradox in Brazil"
subtitle: "TidyTuesday 2026 · Week 4 (Extended) · Brazilian Companies (CNPJ)"
author: "Zhijun He"
date: "2026-02-25"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
    fig-dpi: 180
    fig-align: center
execute:
  warning: false
  message: false
---

## Background

The previous analysis of this dataset established a macro-level capital paradox: the Brazilian
formal sector exhibits extreme capital concentration, with Gini coefficients above 0.98 in every
size tier. That finding tells us the *shape* of inequality but not its *sources*. It also treats
each size category as internally homogeneous, which it is not.

Brazil's official firm-size classification (Simples Nacional) is based on **annual gross revenue**,
not capital stock. A sole proprietor earning 360,000 BRL per year is legally a "micro-enterprise"
regardless of whether they hold 180,000 BRL or 10,000,000 BRL in declared capital stock. This
classification logic creates a natural experiment embedded in the data: firms within the same
regulatory category can have wildly different capital endowments. The spread *within* size tiers
is as analytically interesting as the gaps *between* them.

A second underexplored dimension is governance. The `owner_qualification` field records the
legal role of the controlling person: a Managing Partner who actively runs the business, a Beneficial
Owner who holds equity but may be a passive investor or part of a holding-company chain, or a
solo Entrepreneur operating under their own identity. These governance archetypes map onto well-
established predictions from agency theory: passive beneficial ownership is associated with access
to external capital, while sole entrepreneurs face the highest information asymmetries with lenders.

## Research Questions

Three analytical questions guide this extended analysis:

1. **Within-tier inequality:** Are micro-enterprises themselves internally unequal in capital
   endowments, or is the category relatively flat?
2. **Governance as a capital predictor:** Does owner qualification predict capital accumulation
   *independently* of firm size, and by how much?
3. **The "capital-rich micro":** Among micro-enterprises with unusually high capital stock, what
   governance and legal-form features distinguish them from the capital-poor majority?

## Design Conventions

```{r}
#| label: setup-theme

library(tidyverse)
library(scales)
library(ggridges)
library(patchwork)

# ── Shared palette ───────────────────────────────────────────────────────────
pal_size <- c(
  "Micro"        = "#c3553a",   # terracotta
  "Small"        = "#d9a441",   # amber
  "Medium/Large" = "#2a5a6b"    # deep teal
)

pal_gov <- c(
  "Sole Entrepreneur"                    = "#d9a441",   # amber (solo)
  "Managing Partner /\nAdministrator"    = "#5b8a9a",   # slate blue (active)
  "Beneficial Owner\n(passive/investor)" = "#1d3557"    # deep navy (passive)
)

# ── Shared theme ─────────────────────────────────────────────────────────────
theme_cwc <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title        = element_text(face = "bold", size = base_size + 2,
                                       colour = "#1d3557", margin = margin(b = 4)),
      plot.subtitle     = element_text(colour = "grey35", size = base_size - 2,
                                       lineheight = 1.35, margin = margin(b = 18)),
      plot.caption      = element_text(colour = "grey55", size = base_size - 5,
                                       hjust = 0, margin = margin(t = 12)),
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      strip.text        = element_text(face = "bold", size = base_size - 1,
                                       colour = "#1d3557"),
      strip.background  = element_rect(fill = "grey96", colour = NA),
      panel.grid.minor  = element_blank(),
      panel.grid.major  = element_line(colour = "grey92", linewidth = 0.3),
      panel.spacing     = unit(1.4, "lines"),
      legend.position   = "top",
      legend.title      = element_text(face = "bold", size = base_size - 2),
      legend.text       = element_text(size = base_size - 2),
      axis.title        = element_text(colour = "grey30", size = base_size - 2),
      axis.text         = element_text(colour = "grey40", size = base_size - 3)
    )
}

caption_text <- "Source: Brazilian CNPJ Registry · TidyTuesday 2026 Wk 4 (Extended) · Zhijun He · 2026-02-25"
```

## Data Preparation

```{r}
#| label: data-prep

companies <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv",
  show_col_types = FALSE
)

co <- companies |>
  filter(!is.na(company_size), !is.na(capital_stock), !is.na(owner_qualification),
         capital_stock > 0) |>
  mutate(
    size = case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE                               ~ "Medium/Large"
    ),
    size = factor(size, levels = c("Micro", "Small", "Medium/Large")),

    governance = case_when(
      str_detect(owner_qualification, regex("beneficial owner", ignore_case = TRUE))         ~ "Beneficial Owner\n(passive/investor)",
      str_detect(owner_qualification, regex("entrepreneur|business owner", ignore_case = TRUE)) ~ "Sole Entrepreneur",
      str_detect(owner_qualification, regex("managing partner|administrator", ignore_case = TRUE)) ~ "Managing Partner /\nAdministrator",
      TRUE ~ "Other"
    ),
    governance = factor(governance, levels = c(
      "Sole Entrepreneur",
      "Managing Partner /\nAdministrator",
      "Beneficial Owner\n(passive/investor)",
      "Other"
    )),

    log_capital = log10(capital_stock),

    capital_q75_micro = quantile(capital_stock[size == "Micro"], 0.75, na.rm = TRUE),
    is_capital_rich_micro = (size == "Micro") & (capital_stock >= capital_q75_micro)
  )

cat("Observations:", comma(nrow(co)), "\n")
cat("Governance archetypes:\n"); print(count(co, governance, sort = TRUE))
cat("\nCapital-rich micro threshold: BRL ",
    comma(unique(co$capital_q75_micro[!is.na(co$capital_q75_micro)])[1]), "\n", sep = "")
```

## Visualization 1 — The Power-Law Signature of Brazilian Capital

A plain Lorenz curve compresses the story because every tier is extremely concentrated — the
three Ginis (all above 0.98) plot as almost the same curve. To read the structure honestly, we
need two views: a log-rank concentration curve that separates the tiers where the action lives,
and a direct decomposition of who sits at the top.

```{r}
#| label: fig-lorenz-power
#| fig-width: 12
#| fig-height: 5.8

# ── Gini helper ─────────────────────────────────────────────────────────────
gini <- function(x) {
  x <- sort(x[!is.na(x) & x > 0])
  n <- length(x)
  2 * sum((seq_len(n) / n) * (x / sum(x))) - (n + 1) / n
}

# Compute Gini + Top 1% share per tier; burn both into the legend label
tier_stats <- co |>
  group_by(size) |>
  summarise(
    g    = gini(capital_stock),
    top1 = sum(sort(capital_stock, decreasing = TRUE)[1:ceiling(n() * 0.01)]) /
           sum(capital_stock),
    .groups = "drop"
  ) |>
  mutate(
    label = sprintf("%-13s   Gini %.3f    Top 1%% holds %s",
                    size, g, percent(top1, accuracy = 1))
  )

# ── Panel A: Concentration curve on log-x, zero in-plot annotations ─────────
conc_curve <- co |>
  group_by(size) |>
  arrange(desc(capital_stock), .by_group = TRUE) |>
  mutate(
    rank_pct = row_number() / n(),
    cum_top  = cumsum(capital_stock) / sum(capital_stock)
  ) |>
  ungroup() |>
  filter(rank_pct <= 0.5)

p_conc <- ggplot(conc_curve, aes(x = rank_pct, y = cum_top, colour = size)) +
  annotate("segment", x = 0.01, xend = 0.01, y = 0, yend = 1,
           colour = "grey75", linetype = "dashed", linewidth = 0.45) +
  annotate("text", x = 0.01, y = 1.04, label = "Top 1%",
           colour = "grey40", size = 3.3, fontface = "italic") +
  geom_line(linewidth = 1.5, alpha = 0.92) +
  scale_x_log10(
    breaks = c(0.001, 0.01, 0.1, 0.5),
    labels = c("Top 0.1%", "Top 1%", "Top 10%", "Top 50%"),
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  scale_y_continuous(
    labels = percent_format(),
    limits = c(0, 1.08),
    expand = c(0, 0),
    name   = "Cumulative share of tier's capital"
  ) +
  scale_colour_manual(values = pal_size,
                      labels = tier_stats$label,
                      name   = NULL) +
  labs(
    x = "Rank of firms (log scale, highest capital first) →",
    subtitle = "(A)  Concentration curve — log-x rank reveals the separation\n        between tiers that linear Lorenz cannot show"
  ) +
  theme_cwc() +
  theme(
    legend.position      = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.background    = element_rect(fill = alpha("white", 0.9), colour = NA),
    legend.key.width     = unit(1.5, "lines"),
    legend.text          = element_text(family = "mono", size = 10),
    plot.subtitle        = element_text(colour = "grey35", size = 11,
                                        lineheight = 1.3, margin = margin(b = 10))
  )

# ── Panel B: Stacked decomposition by rank bucket ───────────────────────────
decomp <- co |>
  group_by(size) |>
  arrange(desc(capital_stock), .by_group = TRUE) |>
  mutate(
    rank_pct = row_number() / n(),
    segment  = case_when(
      rank_pct <= 0.01 ~ "Top 1%",
      rank_pct <= 0.10 ~ "Next 9%",
      rank_pct <= 0.50 ~ "Next 40%",
      TRUE             ~ "Bottom 50%"
    ),
    segment  = factor(segment,
                      levels = c("Bottom 50%", "Next 40%", "Next 9%", "Top 1%"))
  ) |>
  group_by(size, segment) |>
  summarise(cap = sum(capital_stock), .groups = "drop") |>
  group_by(size) |>
  mutate(share = cap / sum(cap)) |>
  ungroup()

pal_segment <- c(
  "Top 1%"     = "#1d3557",
  "Next 9%"    = "#457b9d",
  "Next 40%"   = "#a8dadc",
  "Bottom 50%" = "#e9ecef"
)

p_decomp <- ggplot(decomp, aes(x = size, y = share, fill = segment)) +
  geom_col(width = 0.62, colour = "white", linewidth = 1.4) +
  geom_text(
    data = filter(decomp, share >= 0.04),   # suppress tiny stray labels
    aes(label = percent(share, accuracy = 1),
        colour = segment %in% c("Top 1%", "Next 9%")),
    position = position_stack(vjust = 0.5),
    fontface = "bold", size = 4, show.legend = FALSE
  ) +
  scale_fill_manual(
    values = pal_segment,
    breaks = c("Top 1%", "Next 9%", "Next 40%", "Bottom 50%"),
    name   = NULL
  ) +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey25")) +
  scale_y_continuous(labels = percent_format(),
                     expand = c(0, 0),
                     name   = "Share of tier's total capital") +
  labs(
    x = NULL,
    subtitle = "(B)  In every tier, the top 1% of firms holds ~96–99%\n        of the tier's capital. The bottom half is a sliver."
  ) +
  theme_cwc() +
  theme(
    legend.position    = "top",
    panel.grid.major.x = element_blank(),
    plot.subtitle      = element_text(colour = "grey35", size = 11,
                                      lineheight = 1.3, margin = margin(b = 10))
  )

# ── Combine ─────────────────────────────────────────────────────────────────
(p_conc | p_decomp) +
  plot_layout(widths = c(1.15, 1)) +
  plot_annotation(
    title    = "Capital concentration is near-total in every Brazilian firm-size tier",
    subtitle = "Gini coefficients exceed 0.98 across Micro, Small, and Medium/Large — 'micro-enterprise' is not uniquely unequal.\nThe capital trap is a universal power-law structure, not a feature of small firms.",
    caption  = caption_text,
    theme    = theme_cwc()
  ) &
  theme(plot.title = element_text(face = "bold", size = 16, colour = "#1d3557"))
```

Panel A uses a log-scaled rank axis to surface a separation that linear Lorenz curves flatten
away. The legend carries the precise numbers: each tier's Gini and its top-1% capital share.
Panel B decomposes the same fact along a cleaner axis — every tier's top 1% of firms holds
roughly 96–99% of that tier's capital, while the bottom half is a sliver barely visible at the
top of each bar. The "capital trap" described in earlier work is not a peculiarity of
micro-enterprises. It is a **universal power-law structure** across the Brazilian formal sector.
This reframes the rest of the analysis: what differentiates tiers is not their inequality, but
the *who* at the top — the question governance can answer.

## Visualization 2 — Governance as a Distributional Shape, Not Just a Level

```{r}
#| label: fig-governance-ridges
#| fig-width: 12
#| fig-height: 6.2

gov_stats <- co |>
  filter(governance != "Other") |>
  group_by(size, governance) |>
  summarise(
    median_cap = median(capital_stock),
    mean_cap   = mean(capital_stock),
    p90_cap    = quantile(capital_stock, 0.90),
    n          = n(),
    .groups    = "drop"
  ) |>
  mutate(
    med_lab = paste0("median ", number(median_cap, scale = 1e-3, suffix = "k",
                                       big.mark = ",", accuracy = 1))
  )

ggplot(
  co |> filter(governance != "Other"),
  aes(x = log_capital, y = governance, fill = governance)
) +
  geom_density_ridges(
    alpha = 0.85, scale = 0.92, rel_min_height = 0.01,
    quantile_lines = TRUE, quantiles = 2,
    colour = "white", linewidth = 0.45
  ) +
  # mean marker (red tick) — surfaces the heavy-tail of Beneficial Owners
  geom_segment(
    data = gov_stats,
    aes(x = log10(mean_cap), xend = log10(mean_cap),
        y = as.numeric(governance) - 0.02,
        yend = as.numeric(governance) + 0.38),
    inherit.aes = FALSE,
    colour = "#c3553a", linewidth = 0.8
  ) +
  # median label placed above the ridge
  geom_text(
    data = gov_stats,
    aes(x = log10(median_cap), y = as.numeric(governance) + 0.72,
        label = med_lab),
    inherit.aes = FALSE,
    size = 3.0, colour = "grey25", hjust = 0.5
  ) +
  facet_wrap(~size, nrow = 1) +
  scale_fill_manual(values = pal_gov, guide = "none") +
  scale_x_continuous(
    breaks = c(4, 5, 6, 7, 8),
    labels = c("10k", "100k", "1M", "10M", "100M"),
    name   = "Capital stock (BRL, log scale)",
    expand = expansion(mult = c(0.02, 0.04))
  ) +
  scale_y_discrete(expand = expansion(add = c(0.3, 1.1))) +
  labs(
    title    = "Governance shapes the distribution — Managing Partners lead the median, Beneficial Owners carry the tail",
    subtitle = "Density ridges show capital-stock distribution by owner archetype within each size tier.\nWhite line = median  •  Red tick = mean  •  X-axis: log₁₀ scale",
    y        = NULL,
    caption  = caption_text
  ) +
  theme_cwc() +
  theme(panel.grid.major.y = element_blank())
```

The refined picture flips the original interpretation. Looking at medians:

- **Managing Partners** hold the highest median capital, and — crucially — they are the only
  archetype whose median *scales with firm size*: 300k at Micro, 360k at Small, **1,000k at
  Medium/Large**. Active operators accumulate capital as they grow.
- **Sole Entrepreneurs** show a flat median (~300k) that barely moves across tiers — consistent
  with the information-asymmetry prediction from agency theory.
- **Beneficial Owners** are the most distinctive: their median is low and flat (~300k across all
  three tiers), but the mean (red tick) sits far to the right of the median, and the distribution
  has a visible long right tail. This is a **heavy-tail archetype** — most BO firms are small,
  but a minority hold enormous capital. That shape is the statistical fingerprint of
  holding-company architecture: passive ownership is rare, and when it appears, it spans the
  full range from shell companies to major investment vehicles.

The governance premium, then, is not uniform. It shows up as a *level premium* for Managing
Partners (especially in larger tiers) and as a *variance premium* for Beneficial Owners. Policies
that assume a single governance hierarchy miss this distributional shape.

## Visualization 3 — The Escape Route Is a Partnership, Not an LLC

```{r}
#| label: fig-escape
#| fig-width: 11
#| fig-height: 6

micro_profile <- co |>
  filter(size == "Micro", governance != "Other") |>
  mutate(
    legal_short = case_when(
      str_detect(legal_nature, "LLC")                 ~ "LLC",
      str_detect(legal_nature, "Sole Proprietorship") ~ "Sole\nProprietary",
      str_detect(legal_nature, "Individual Limited")  ~ "EIRELI",
      str_detect(legal_nature, "Simple")              ~ "Simple\nPartnership",
      str_detect(legal_nature, "Silent")              ~ "Silent\nPartnership",
      str_detect(legal_nature, "Corporation")         ~ "Corporation",
      TRUE                                            ~ "Other"
    )
  )

escape_rate <- micro_profile |>
  group_by(legal_short, governance) |>
  summarise(
    n_total    = n(),
    n_rich     = sum(is_capital_rich_micro),
    escape_pct = n_rich / n_total,
    .groups    = "drop"
  ) |>
  filter(n_total >= 20) |>
  mutate(
    combo_label = paste0(legal_short, " × ", str_replace(governance, "\\n", " "))
  ) |>
  arrange(desc(escape_pct))

ggplot(
  escape_rate,
  aes(x = fct_reorder(combo_label, escape_pct),
      y = escape_pct,
      fill = governance)
) +
  geom_col(width = 0.7, alpha = 0.95) +
  geom_text(aes(label = percent(escape_pct, accuracy = 0.1)),
            hjust = -0.18, size = 3.3, colour = "grey20") +
  geom_text(aes(label = paste0("n = ", comma(n_total)), y = 0.003),
            hjust = 0, size = 2.8, colour = "white", fontface = "bold") +
  coord_flip(clip = "off") +
  scale_y_continuous(
    labels = percent_format(),
    limits = c(0, 0.58),
    expand = expansion(mult = c(0, 0.05)),
    name   = "Share of cell above the 75th capital percentile (BRL 500,000)"
  ) +
  scale_fill_manual(values = pal_gov, name = "Governance archetype",
                    guide = guide_legend(reverse = TRUE)) +
  labs(
    title    = "The micro-enterprise escape profile is a partnership with a passive equity holder",
    subtitle = "Share of capital-rich micros (>BRL 500k) by legal form × governance cell.\nCells with fewer than 20 firms excluded. Counts shown in white.",
    x        = NULL,
    caption  = caption_text
  ) +
  theme_cwc() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y        = element_text(size = 10, colour = "grey20"),
    plot.margin        = margin(10, 40, 10, 10)
  )
```

Ranking the legal-form × governance cells that survive the n ≥ 20 cutoff:

- **Simple Partnership × Beneficial Owner** (48.2%) — almost half of micros in this cell are
  capital-rich.
- **Simple Partnership × Managing Partner** (41.6%) — second highest.
- Sole Proprietary × Sole Entrepreneur (34.3%), LLC × Managing Partner (29.0%), EIRELI ×
  Beneficial Owner (24.1%) trail behind.

The pattern is not about LLCs as a legal form — it is about **multi-party partnership
structures**. The Simple Partnership requires two or more parties on paper; when a Beneficial
Owner sits behind them, the cell's escape rate nearly doubles the base rate. This is the
fingerprint of holding-company architecture operating through partnership law: a passive equity
holder channelling capital into a nominally two-party vehicle whose revenue happens to sit below
the Simples threshold.

EIRELI (Brazil's individual limited liability form, created for single-person companies to
separate personal from business liability) appears further down the list. The popular framing
of "EIRELI as escape route" is not what the data show. The escape route is **partnership law
plus passive ownership**.

## Synthesis

This extended analysis reframes the capital trap finding in three directions.

**First**, capital concentration is *universal* across the Brazilian formal sector, not a
micro-enterprise pathology. Gini coefficients above 0.98 in every size tier — and top-1% capital
shares of 96–99% — mean that treating "micro-enterprise" as a uniquely unequal category misreads
the structure. The same power-law sits behind every regulatory label.

**Second**, governance does not deliver a single capital hierarchy. Managing Partners lead at
the median and scale with firm size. Sole Entrepreneurs stay flat. Beneficial Owners are a
heavy-tail archetype — most are small, a few are very large. Size and governance are orthogonal
dimensions, but they interact differently depending on which moment of the distribution
(median vs. tail) one cares about.

**Third**, the capital-rich micro profile concentrates in **Simple Partnership structures under
Beneficial Ownership** — a signature consistent with holding-company architecture rather than
organic small-business growth. The policy implication is direct: if Brazil's development goal
is to support genuinely capital-constrained entrepreneurs, the targeting instrument must
incorporate legal form and governance alongside revenue thresholds. Revenue alone is porous —
partnership law plus passive ownership is the channel through which capital-rich firms wear
micro-enterprise clothing.
