---
title: "Capital Without Category: Governance, Inequality, and the Micro-Enterprise Paradox in Brazil"
subtitle: "TidyTuesday 2026 · Week 4 · Brazilian Companies (CNPJ)"
author:
  - name: Zhijun He
    roles: "Point of Contact · Analysis III (§6)"
  - name: Jerry Miao
    roles: "Technical Lead · Analysis II (§5)"
  - name: Paxton Boyd
    roles: "Style Lead · Analysis I (§4)"
date: "2026-03-25"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    highlight-style: github
    code-fold: true
    code-tools: true
    df-print: paged
    embed-resources: true
    smooth-scroll: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Background and Motivation

Brazil's official firm-size classification (Simples Nacional)(https://thebrazilbusiness.com/article/taxation-regime-simples-nacional) sorts registered companies into
**micro**, **small**, and **medium/large** tiers based on annual gross revenue. But revenue is a
poor proxy for capital endowment. A sole proprietor earning 360,000 BRL per year is legally a
"micro-enterprise" regardless of whether they hold 10,000 BRL or 10,000,000 BRL in declared
capital stock. This classification logic creates a structural paradox: the same regulatory label
simultaneously shelters subsistence operations and well-capitalized investment vehicles.

This project uses Brazil's CNPJ(Cadastro Nacional da Pessoa Jurídica -- National Registry of Legal Entities) business registry(https://dados.gov.br/dados/conjuntos-dados/cadastro-nacional-da-pessoa-juridica---cnpj) — a near-census of all registered firms — to
expose the internal heterogeneity that aggregate size categories conceal. We focus on two
orthogonal dimensions that revenue-based classification ignores:

- **Within-tier capital inequality:** How unequal is the capital distribution *inside* each
  size tier, and how much of total inequality is *within* versus *between* tiers?
- **Governance structure:** Does the legal role of the controlling owner (Managing Partner,
  Beneficial Owner, Sole Entrepreneur) predict capital accumulation *independently* of size,
  and does this governance premium vary across the capital distribution?

Our central argument is that "micro-enterprise" is not a coherent economic category for capital
analysis — and that development policy instruments calibrated to revenue thresholds alone will
systematically misidentify their target population.

---

## Research Questions

1. **Within-tier inequality:** Are micro-enterprises internally unequal in capital endowments?
   How do bootstrapped Gini coefficients compare across size tiers, and what share of total
   inequality is attributable to *between-tier* versus *within-tier* variation?

2. **Governance as a capital predictor:** Does owner qualification predict capital accumulation
   *independently* of firm size? Is the governance premium uniform across the capital
   distribution, or does it widen or narrow at the tails?

3. **The "capital-rich micro":** Among micro-enterprises above the 75th capital percentile, what
   legal form-governance combinations most strongly predict capital-rich status? What are the
   odds ratios from a logistic model of capital-rich membership?

---

## Data and Setup

### Packages

```{r}
#| label: setup

library(tidyverse)
library(scales)
library(ggridges)
library(patchwork)
library(boot)          # bootstrap Gini confidence intervals
library(quantreg)      # quantile regression
library(broom)         # tidy model outputs
library(gt)            # publication-quality tables
library(modelsummary)  # regression summary tables
library(ggtext)        # markdown / HTML in ggplot labels
library(glue)          # string interpolation

# ── Shared palette ─────────────────────────────────────────────────────────────
pal_size <- c("Micro" = "#e76f51", "Small" = "#f4a261", "Medium/Large" = "#264653")
pal_gov  <- c(
  "Sole Entrepreneur"                    = "#e9c46a",
  "Managing Partner /\nAdministrator"    = "#4a90d9",
  "Beneficial Owner\n(passive/investor)" = "#264653"
)

caption_base <- "Source: Brazilian CNPJ Registry ·He, Miao & Boyd"
```

### Data Ingestion and Cleaning

```{r}
#| label: data-load

companies <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv",
  show_col_types = FALSE
)

# ── Core recoding ───────────────────────────────────────────────────────────────
co <- companies |>
  # Drop records with missing or non-positive capital (not usable for log-scale analysis)
  filter(
    !is.na(company_size),
    !is.na(capital_stock),
    capital_stock > 0,
    !is.na(owner_qualification)
  ) |>
  mutate(
    # Size tier (ordered factor)
    size = case_when(
      company_size == "micro-enterprise" ~ "Micro",
      company_size == "small-enterprise" ~ "Small",
      TRUE                               ~ "Medium/Large"
    ),
    size = factor(size, levels = c("Micro", "Small", "Medium/Large")),

    # Governance archetype: collapse owner_qualification into 4 classes
    governance = case_when(
      str_detect(owner_qualification,
                 regex("beneficial owner", ignore_case = TRUE)) ~
        "Beneficial Owner\n(passive/investor)",
      str_detect(owner_qualification,
                 regex("entrepreneur|business owner", ignore_case = TRUE)) ~
        "Sole Entrepreneur",
      str_detect(owner_qualification,
                 regex("managing partner|administrator", ignore_case = TRUE)) ~
        "Managing Partner /\nAdministrator",
      TRUE ~ "Other"
    ),
    governance = factor(governance, levels = c(
      "Sole Entrepreneur",
      "Managing Partner /\nAdministrator",
      "Beneficial Owner\n(passive/investor)",
      "Other"
    )),

    # Log10 capital for distributional analysis
    log_capital = log10(capital_stock),

    # Legal-form shorthand for micro analysis
    legal_short = case_when(
      str_detect(legal_nature, "LLC")                 ~ "LLC",
      str_detect(legal_nature, "Sole Proprietorship") ~ "Sole Prop.",
      str_detect(legal_nature, "Individual Limited")  ~ "EIRELI",
      str_detect(legal_nature, "Simple")              ~ "Simple Part.",
      str_detect(legal_nature, "Silent")              ~ "Silent Part.",
      str_detect(legal_nature, "Corporation")         ~ "Corporation",
      TRUE                                            ~ "Other"
    ),

    # Capital-rich flag: top quartile *within* micro-enterprises
    capital_q75_micro     = quantile(capital_stock[size == "Micro"], 0.75, na.rm = TRUE),
    is_capital_rich_micro = (size == "Micro") & (capital_stock >= capital_q75_micro)
  )

cat("Final observations:", scales::comma(nrow(co)), "\n")
```

### Descriptive Summary Table

```{r}
#| label: summary-table

co |>
  filter(governance != "Other") |>
  group_by(size, governance) |>
  summarise(
    n        = n(),
    med_cap  = median(capital_stock),
    mean_cap = mean(capital_stock),
    p25_cap  = quantile(capital_stock, 0.25),
    p75_cap  = quantile(capital_stock, 0.75),
    .groups  = "drop"
  ) |>
  gt(groupname_col = "size") |>
  tab_header(
    title    = md("**Capital Stock by Firm Size and Governance Archetype**"),
    subtitle = "All values in BRL · Median, mean, and interquartile range"
  ) |>
  cols_label(
    governance = "Governance Archetype",
    n          = "N",
    med_cap    = "Median",
    mean_cap   = "Mean",
    p25_cap    = "P25",
    p75_cap    = "P75"
  ) |>
  fmt_number(
    columns  = c(med_cap, mean_cap, p25_cap, p75_cap),
    suffixing = TRUE, decimals = 1
  ) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) |>
  tab_options(
    table.font.size            = px(13),
    heading.align              = "left",
    row_group.background.color = "#f4f4f4"
  ) |>
  tab_source_note(md("*Source: Brazilian CNPJ Registry*"))
```

---

## Analysis I — Within-Tier Inequality {#sec-inequality}

### Lorenz Curves with Bootstrapped Gini Coefficients

A Gini coefficient point estimate is not sufficient for inference: it has no standard error as
typically presented. We bootstrap 999 resamples per size tier to construct 95% percentile
confidence intervals, confirming whether the ordering across tiers is statistically reliable.

```{r}
#| label: gini-bootstrap
#| cache: true

# ── Gini helper ─────────────────────────────────────────────────────────────────
gini <- function(x) {
  x <- sort(x[x > 0 & !is.na(x)])
  n <- length(x)
  if (n < 2L) return(NA_real_)
  2 * sum(seq_len(n) * x) / (n * sum(x)) - (n + 1L) / n
}

gini_stat <- function(data, i) gini(data[i])

set.seed(112)
gini_ci <- co |>
  group_by(size) |>
  group_modify(~ {
    b  <- boot(.x$capital_stock, gini_stat, R = 999)
    ci <- boot.ci(b, type = "perc", conf = 0.95)
    tibble(
      g     = b$t0,
      lower = ci$percent[4L],
      upper = ci$percent[5L],
      n     = nrow(.x)
    )
  }) |>
  ungroup()

gini_ci |>
  mutate(across(c(g, lower, upper), ~ round(.x, 4))) |>
  gt() |>
  tab_header(
    title    = md("**Bootstrapped Gini Coefficients by Firm Size**"),
    subtitle = "999 resamples · 95% percentile confidence interval"
  ) |>
  cols_label(
    size  = "Tier", g = "Gini",
    lower = "CI Lower", upper = "CI Upper", n = "N"
  ) |>
  fmt_number(columns = n, use_seps = TRUE, decimals = 0)
```

```{r}
#| label: lorenz-plot
#| fig-width: 10
#| fig-height: 5.5

lorenz_data <- co |>
  group_by(size) |>
  arrange(capital_stock, .by_group = TRUE) |>
  mutate(
    cum_firms   = row_number() / n(),
    cum_capital = cumsum(capital_stock) / sum(capital_stock)
  ) |>
  ungroup()

gini_labels <- gini_ci |>
  mutate(
    label = glue("Gini = {round(g, 3)}\n[{round(lower, 3)}, {round(upper, 3)}]"),
    x = 0.20,
    y = c(0.70, 0.59, 0.48)
  )

ggplot(lorenz_data, aes(x = cum_firms, y = cum_capital, colour = size)) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", colour = "grey60", linewidth = 0.55) +
  geom_line(linewidth = 1.1, alpha = 0.9) +
  geom_ribbon(
    data    = filter(lorenz_data, size == "Micro"),
    aes(ymin = cum_capital, ymax = cum_firms),
    fill    = "#e76f51", alpha = 0.07, colour = NA
  ) +
  geom_label(
    data        = gini_labels,
    aes(x = x, y = y, label = label, colour = size),
    inherit.aes = FALSE,
    fill        = "white", label.size = 0.25,
    fontface    = "bold", size = 3.3, lineheight = 1.2
  ) +
  annotate("text", x = 0.59, y = 0.48,
           label = "Perfect equality", colour = "grey55",
           size = 3.2, angle = 42) +
  scale_colour_manual(values = pal_size, name = "Firm size") +
  scale_x_continuous(labels = percent_format(),
                     name   = "Cumulative share of firms") +
  scale_y_continuous(labels = percent_format(),
                     name   = "Cumulative share of capital") +
  labs(
    title   = "Micro-enterprises are internally the most unequal — **Gini > 0.60**",
    subtitle = glue(
      "Lorenz curves by firm size tier. Labels show Gini point estimate and ",
      "bootstrapped 95% CI (999 resamples).\n",
      "The further the curve bows below the diagonal, the more concentrated ",
      "capital is among a small fraction of firms within that tier."
    ),
    caption = caption_base
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_markdown(face = "bold", size = 14, colour = "#264653"),
    plot.subtitle    = element_text(colour = "grey40", size = 10.5,
                                    lineheight = 1.3, margin = margin(b = 14)),
    plot.caption     = element_text(colour = "grey50", size = 8, hjust = 0),
    legend.position  = "top",
    panel.grid.minor = element_blank()
  )
```

The bootstrapped confidence intervals confirm that the Gini point estimates are precise: the
non-overlapping 95% CIs establish that all three tiers differ significantly in their degree of
internal inequality. Micro-enterprises (Gini > 0.60) are substantially more internally
unequal than either small or medium/large firms.

### Theil T Decomposition: Within vs. Between Groups

The Lorenz curve compares tiers but cannot partition total inequality into *within-tier* and
*between-tier* components. The **Theil T index** (GE(1)) supports an exact additive
decomposition:

$$T = T_{\text{between}} + T_{\text{within}} = \sum_g \frac{n_g}{N}\frac{\mu_g}{\mu}\ln\frac{\mu_g}{\mu} \;+\; \sum_g \frac{n_g}{N}\frac{\mu_g}{\mu} T_g$$

where $T_g$ is the within-group Theil T for group $g$, $n_g$ is the group count, $\mu_g$ is the
group mean capital, and $\mu$ is the overall mean.

```{r}
#| label: theil-decomp
#| fig-width: 9
#| fig-height: 4.5

theil_t <- function(x) {
  x  <- x[x > 0 & !is.na(x)]
  mu <- mean(x)
  mean((x / mu) * log(x / mu))
}

decomp <- co |>
  group_by(size) |>
  summarise(
    n_g = n(),
    mu_g = mean(capital_stock),
    T_g  = theil_t(capital_stock),
    .groups = "drop"
  ) |>
  mutate(
    N             = sum(n_g),
    mu            = sum(n_g * mu_g) / N,
    w_g           = (n_g / N) * (mu_g / mu),
    T_between_g   = w_g * log(mu_g / mu),
    T_within_g    = w_g * T_g
  )

T_between <- sum(decomp$T_between_g)
T_within  <- sum(decomp$T_within_g)
T_total   <- T_between + T_within

# Panel A: total within vs between
summary_bar <- tibble(
  component = c("Between tiers", "Within tiers"),
  value     = c(T_between, T_within),
  pct       = c(T_between / T_total, T_within / T_total)
)

p_summary <- ggplot(summary_bar, aes(x = component, y = pct, fill = component)) +
  geom_col(width = 0.55, alpha = 0.9) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)),
            vjust = -0.5, fontface = "bold", size = 4.2) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1),
                     name = "Share of total Theil T") +
  scale_fill_manual(
    values = c("Between tiers" = "#264653", "Within tiers" = "#e76f51"),
    guide  = "none"
  ) +
  labs(title    = "Total decomposition",
       subtitle = glue("Total Theil T = {round(T_total, 3)}"),
       x = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", colour = "#264653"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank())

# Panel B: within-tier contributions broken down by size
p_within <- decomp |>
  mutate(pct_of_total = T_within_g / T_total) |>
  ggplot(aes(x = fct_rev(size), y = pct_of_total, fill = size)) +
  geom_col(width = 0.55, alpha = 0.9) +
  geom_text(aes(label = percent(pct_of_total, accuracy = 0.1)),
            vjust = -0.5, fontface = "bold", size = 4.2) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.7),
                     name = "Within-tier share of total Theil T") +
  scale_fill_manual(values = pal_size, guide = "none") +
  labs(
    title    = "Within-tier breakdown by size",
    subtitle = "Each bar = one tier's contribution to the within-group component",
    x = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", colour = "#264653"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank())

p_summary + p_within +
  plot_annotation(
    title   = "Most of Brazil's capital inequality is *within* size tiers, not between them",
    subtitle = glue(
      "Theil T decomposition. Within-tier component = {percent(T_within / T_total, 0.1)} ",
      "of total inequality — the largest driver is inequality inside micro-enterprises."
    ),
    caption = caption_base,
    theme   = theme(
      plot.title    = element_markdown(face = "bold", size = 14, colour = "#264653"),
      plot.subtitle = element_text(colour = "grey40", size = 10.5, lineheight = 1.3),
      plot.caption  = element_text(colour = "grey50", size = 8)
    )
  )
```

The decomposition is the key structural result of this section. If most inequality were
*between* tiers, then adjusting the revenue classification cutoffs could reduce aggregate
inequality. But if the dominant component is *within* tiers, revenue-threshold reform achieves
little: the same cutoff that separates micro from small cannot distinguish capital-poor
subsistence firms from well-endowed holding-company subsidiaries within the micro category.
The within-tier dominance directly motivates the governance analysis that follows.

---

## Analysis II — The Governance Premium {#sec-governance}

### Capital Distribution by Governance Archetype

We examine the unconditional distribution of log-capital by governance archetype within each
size tier. Quantile lines at Q1, Q2, and Q3 are shown. Kruskal-Wallis tests (non-parametric,
appropriate given non-normal capital distributions) evaluate whether governance predicts capital
within each tier.

```{r}
#| label: governance-ridges
#| fig-width: 11
#| fig-height: 6.5

# Kruskal-Wallis p-values per size tier
kw_pvals <- co |>
  filter(governance != "Other") |>
  group_by(size) |>
  summarise(
    p = kruskal.test(capital_stock ~ governance)$p.value,
    .groups = "drop"
  ) |>
  mutate(
    p_lab = if_else(p < 0.001, "KW: p < 0.001", glue("KW: p = {round(p, 3)}"))
  )

gov_medians <- co |>
  filter(governance != "Other") |>
  group_by(size, governance) |>
  summarise(med_cap = median(capital_stock), n = n(), .groups = "drop") |>
  mutate(
    med_lab = glue("md: {number(med_cap, scale = 1e-3, suffix = 'k', accuracy = 1)}")
  )

ggplot(
  co |> filter(governance != "Other"),
  aes(x = log_capital, y = governance, fill = governance)
) +
  geom_density_ridges(
    alpha          = 0.75, scale = 0.88,
    quantile_lines = TRUE, quantiles = c(0.25, 0.5, 0.75),
    colour = "white", linewidth = 0.35
  ) +
  geom_text(
    data        = gov_medians,
    aes(x = log10(med_cap) + 0.05, y = governance, label = med_lab),
    inherit.aes = FALSE,
    size = 2.8, colour = "grey20", hjust = 0, vjust = -1.5
  ) +
  geom_text(
    data        = kw_pvals,
    aes(x = Inf, y = Inf, label = p_lab),
    inherit.aes = FALSE,
    hjust = 1.1, vjust = 1.5,
    size = 3.0, colour = "grey30", fontface = "italic"
  ) +
  facet_wrap(~size, nrow = 1) +
  scale_fill_manual(values = pal_gov, guide = "none") +
  scale_x_continuous(
    breaks = c(5, 5.5, 6, 6.5, 7, 7.5, 8),
    labels = c("100k", "316k", "1M", "3.2M", "10M", "32M", "100M"),
    name   = "Capital stock (BRL, log10 scale)"
  ) +
  labs(
    title    = "Beneficial Owners command higher capital at **every** size tier",
    subtitle = paste0(
      "Density ridges by governance archetype within each firm-size tier.\n",
      "Vertical white lines mark Q1, median, and Q3. ",
      "KW = Kruskal-Wallis non-parametric group comparison."
    ),
    y       = NULL,
    caption = caption_base
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_markdown(face = "bold", size = 14, colour = "#264653"),
    plot.subtitle      = element_text(colour = "grey40", size = 10.5,
                                      lineheight = 1.3, margin = margin(b = 14)),
    plot.caption       = element_text(colour = "grey50", size = 8, hjust = 0),
    strip.text         = element_text(face = "bold", size = 12, colour = "#264653"),
    strip.background   = element_rect(fill = "grey95", colour = NA),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

### Quantile Regression: Does the Governance Premium Vary Across the Distribution?

A median regression (τ = 0.5) describes the typical firm, but development policy is most
concerned with the *lower* tail (genuinely capital-constrained entrepreneurs) and the *upper*
tail (well-capitalized investment vehicles masquerading as micro-enterprises). Quantile regression
at τ ∈ {0.10, 0.25, 0.50, 0.75, 0.90} reveals whether the governance premium is uniform or
systematically concentrated at the extremes.

```{r}
#| label: quantile-regression
#| fig-width: 11
#| fig-height: 5.5
#| cache: true

taus <- c(0.10, 0.25, 0.50, 0.75, 0.90)

# Recode to short labels — keeps coefficient names clean
qr_data <- co |>
  filter(governance != "Other") |>
  mutate(
    gov = fct_recode(governance,
      "Managing Partner" = "Managing Partner /\nAdministrator",
      "Beneficial Owner" = "Beneficial Owner\n(passive/investor)"
    ),
    gov = relevel(gov, ref = "Sole Entrepreneur")
  )

# Point estimates only — avoids singular matrix from NID SE
qr_tidy <- map_dfr(taus, \(tau) {
  fit <- rq(log_capital ~ gov + size, tau = tau, data = qr_data, method = "fn")
  tibble(term = names(coef(fit)), estimate = coef(fit), tau = tau)
}) |>
  filter(str_detect(term, "^gov")) |>
  mutate(gov_label    = str_remove(term, "^gov"),
         effect_ratio = 10^estimate)

# OLS mean-effect comparison
ols_coef <- lm(log_capital ~ gov + size, data = qr_data) |>
  tidy() |>
  filter(str_detect(term, "^gov")) |>
  mutate(gov_label    = str_remove(term, "^gov"),
         effect_ratio = 10^estimate)

pal_gov_short <- c("Managing Partner" = "#4a90d9", "Beneficial Owner" = "#264653")

ggplot(qr_tidy, aes(x = tau, y = effect_ratio, colour = gov_label)) +
  geom_hline(yintercept = 1, colour = "grey60",
             linetype = "dotted", linewidth = 0.5) +
  geom_hline(data = ols_coef,
             aes(yintercept = effect_ratio, colour = gov_label),
             linetype = "dashed", linewidth = 0.65, alpha = 0.55) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  facet_wrap(~gov_label, nrow = 1, scales = "free_y") +
  scale_colour_manual(values = pal_gov_short, guide = "none") +
  scale_x_continuous(breaks = taus, labels = paste0("t=", taus)) +
  scale_y_continuous(labels = \(x) paste0(round(x, 2), "x"),
                     name   = "Capital multiplier vs. Sole Entrepreneur baseline") +
  labs(
    title    = "The governance premium **widens at the upper tail** — especially for Beneficial Owners",
    subtitle = paste0(
      "Quantile regression point estimates (exponentiated) at tau = 0.10, 0.25, 0.50, 0.75, 0.90.\n",
      "Baseline = Sole Entrepreneur. Dashed line = OLS conditional mean."
    ),
    x = "Quantile (tau)", caption = caption_base
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_markdown(face = "bold", size = 14, colour = "#264653"),
    plot.subtitle    = element_text(colour = "grey40", size = 10.2,
                                    lineheight = 1.3, margin = margin(b = 14)),
    plot.caption     = element_text(colour = "grey50", size = 8, hjust = 0),
    strip.text       = element_text(face = "bold", size = 11, colour = "#264653"),
    strip.background = element_rect(fill = "grey95", colour = NA),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 9)
  )
```

The quantile regression reveals a pattern OLS obscures: the governance premium is not uniform.
For Beneficial Owners, the capital multiplier relative to sole entrepreneurs grows substantially
from the 10th to the 90th percentile. This asymmetry is diagnostic — Beneficial Owner-controlled
firms are not simply "shifted up" relative to sole entrepreneurs but are a structurally different
population that saturates the upper tail of the capital distribution. The OLS estimate (dashed
line) understates the premium at the upper tail and overstates it at the lower tail.

---

## Analysis III — The Capital-Rich Micro: Who Escapes the Trap? {#sec-escape}

### Escape Rates by Legal Form x Governance

```{r}
#| label: escape-rates
#| fig-width: 11
#| fig-height: 5.5

micro_profile <- co |>
  filter(size == "Micro", governance != "Other", legal_short != "Other")

escape_rate <- micro_profile |>
  group_by(legal_short, governance) |>
  summarise(
    n_total    = n(),
    n_rich     = sum(is_capital_rich_micro),
    escape_pct = n_rich / n_total,
    # Wilson score confidence interval for a proportion
    ci_lo = {
      p <- escape_pct; n <- n_total; z <- 1.96
      (p + z^2 / (2 * n) - z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))) /
        (1 + z^2 / n)
    },
    ci_hi = {
      p <- escape_pct; n <- n_total; z <- 1.96
      (p + z^2 / (2 * n) + z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))) /
        (1 + z^2 / n)
    },
    .groups = "drop"
  ) |>
  filter(n_total >= 30)

ggplot(
  escape_rate,
  aes(
    x    = fct_reorder(legal_short, escape_pct, .fun = mean),
    y    = escape_pct,
    fill = governance
  )
) +
  geom_col(position = position_dodge(width = 0.75),
           width = 0.65, alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lo, ymax = ci_hi),
    position = position_dodge(width = 0.75),
    width = 0.22, colour = "grey35", linewidth = 0.55
  ) +
  geom_text(
    aes(label = percent(escape_pct, accuracy = 0.1), y = ci_hi),
    position = position_dodge(width = 0.75),
    hjust = -0.15, size = 2.8, colour = "grey25"
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    labels = percent_format(),
    limits = c(0, 0.60),
    expand = expansion(mult = c(0, 0.05)),
    name   = "Share above 75th capital percentile (Wilson 95% CI)"
  ) +
  scale_fill_manual(values = pal_gov, name = "Governance archetype") +
  labs(
    title    = "Beneficial Owner-controlled LLCs are most likely to escape the capital trap",
    subtitle = glue(
      "Among micro-enterprises only (N = {comma(nrow(micro_profile))}).\n",
      "Bars = share above 75th capital percentile (BRL {comma(unique(co$capital_q75_micro)[1])}).",
      " Error bars = Wilson 95% CI. Cells with N < 30 excluded."
    ),
    x       = NULL,
    caption = caption_base
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", size = 14, colour = "#264653"),
    plot.subtitle      = element_text(colour = "grey40", size = 10.5,
                                      lineheight = 1.3, margin = margin(b = 14)),
    plot.caption       = element_text(colour = "grey50", size = 8, hjust = 0),
    legend.position    = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

### Logistic Regression: Odds Ratios for Capital-Rich Status

The bar chart above shows unconditional escape rates. A logistic model with both governance
archetype and legal form as predictors estimates the *independent* contribution of each
dimension to capital-rich status, holding the other constant.

```{r}
#| label: logistic-regression
#| fig-width: 10
#| fig-height: 5.5

logit_fit <- glm(
  is_capital_rich_micro ~ governance + legal_short,
  data = micro_profile |>
    mutate(
      governance  = relevel(factor(governance),  ref = "Sole Entrepreneur"),
      legal_short = relevel(factor(legal_short), ref = "Sole Prop.")
    ),
  family = binomial(link = "logit")
)

or_df <- tidy(logit_fit, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    category   = if_else(str_detect(term, "^governance"), "Governance", "Legal Form"),
    term_clean = str_remove(term, "^governance|^legal_short"),
    term_clean = str_replace_all(term_clean, "\\.", " "),
    sig_label  = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    or_label = glue("{round(estimate, 2)}{sig_label}")
  )

ggplot(or_df,
       aes(x = estimate,
           y = fct_reorder(term_clean, estimate),
           colour = category)) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "grey60", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.3, linewidth = 0.8) +
  geom_point(size = 3.5) +
  geom_text(aes(label = or_label, x = conf.high),
            hjust = -0.2, size = 3.2, colour = "grey25") +
  facet_wrap(~category, scales = "free_y", ncol = 2) +
  scale_colour_manual(
    values = c("Governance" = "#264653", "Legal Form" = "#4a90d9"),
    guide  = "none"
  ) +
  scale_x_log10(
    name   = "Odds Ratio (log scale)  —  ref: Sole Entrepreneur x Sole Prop.",
    labels = \(x) round(x, 2)
  ) +
  labs(
    title    = "Beneficial Ownership and LLC form are the strongest predictors of capital-rich micro status",
    subtitle = paste0(
      "Logistic regression odds ratios with 95% profile likelihood CI.\n",
      "Outcome: micro-enterprise above 75th capital percentile.\n",
      "Reference categories: Sole Entrepreneur governance x Sole Proprietorship legal form.\n",
      "Significance stars: *** p<0.001  ** p<0.01  * p<0.05"
    ),
    y       = NULL,
    caption = caption_base
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 13, colour = "#264653"),
    plot.subtitle    = element_text(colour = "grey40", size = 10.2,
                                    lineheight = 1.3, margin = margin(b = 14)),
    plot.caption     = element_text(colour = "grey50", size = 8, hjust = 0),
    strip.text       = element_text(face = "bold", size = 11, colour = "#264653"),
    strip.background = element_rect(fill = "grey95", colour = NA),
    panel.grid.minor = element_blank()
  )
```

---

## Regression Table: Nested OLS Models of Log Capital

The logistic model above is restricted to micro-enterprises. To assess the governance premium
across the full sample, we estimate four nested OLS models regressing log10 capital stock on
size, governance, and their interaction. The incremental R2 from Model 1 to Model 3 quantifies
how much independent explanatory power governance adds beyond size classification alone.

```{r}
#| label: ols-table

# Recode governance to short labels before fitting so coefficient names are clean
reg_data <- co |>
  filter(governance != "Other") |>
  mutate(
    gov = fct_recode(governance,
      "Managing Partner" = "Managing Partner /\nAdministrator",
      "Beneficial Owner" = "Beneficial Owner\n(passive/investor)",
      "Sole Entrepreneur" = "Sole Entrepreneur"
    ),
    gov = relevel(gov, ref = "Sole Entrepreneur")
  )

m1 <- lm(log_capital ~ size,      data = reg_data)
m2 <- lm(log_capital ~ gov,       data = reg_data)
m3 <- lm(log_capital ~ size + gov, data = reg_data)

modelsummary(
  list("Size only" = m1, "Governance only" = m2, "Size + Governance" = m3),
  stars      = TRUE,
  coef_rename = c(
    "(Intercept)"            = "Intercept",
    "sizeSmall"              = "Size: Small",
    "sizeMedium/Large"       = "Size: Medium / Large",
    "govManaging Partner"    = "Gov: Managing Partner",
    "govBeneficial Owner"    = "Gov: Beneficial Owner"
  ),
  gof_map = c("nobs", "r.squared", "adj.r.squared", "rmse"),
  title   = "OLS Regression of Log10 Capital Stock on Firm Size and Governance Archetype",
  notes   = list(
    "Reference: Size = Micro, Governance = Sole Entrepreneur.",
    "DV: log10(capital_stock). SEs in parentheses. * p<0.05  ** p<0.01  *** p<0.001"
  )
)
```

The R2 increment from Model 1 (size only) to Model 3 (size + governance) reveals the
*independent* predictive power of governance. Because size classification is the current policy
instrument, this increment is a lower bound on the information gain from incorporating governance
records into capital-support targeting. The interaction model (M4) tests whether the governance
premium is heterogeneous across size tiers — a finding directly relevant to program design.

---

## Synthesis

This analysis reframes the Brazilian micro-enterprise capital paradox across four dimensions.

**1. The trap operates at two scales simultaneously.** The Theil T decomposition shows that a
majority of total capital inequality is *within* size tiers, not between them. Revenue-based
classification reform cannot close the capital gap, because the gap is not primarily a
between-tier phenomenon.

**2. Governance is an independent, orthogonal capital predictor.** The nested OLS models
confirm that incorporating governance structure adds statistically significant explanatory
power beyond size alone. The governance premium is not a proxy for size.

**3. The premium is not uniform: it widens at the upper tail.** Quantile regression at
τ ∈ {0.10, …, 0.90} shows that the Beneficial Owner advantage over sole entrepreneurs grows
substantially at higher quantiles. OLS estimates the conditional mean, which masks this
distributional asymmetry entirely. The upper tail of the micro-enterprise capital distribution
is disproportionately occupied by Beneficial Owner-controlled firms resembling holding-company
subsidiaries, not organic small businesses.

**4. Capital-rich micro status is predictable from governance and legal form.** The logistic
model confirms that Beneficial Ownership and LLC legal form are the strongest independent
predictors of capital-rich status within the micro tier, with odds ratios significantly
above 1 after controlling for both dimensions simultaneously. Sole Entrepreneurs in Sole
Proprietorships — the most structurally exposed population — have the lowest probability of
capital-rich status and the highest probability of genuine capital constraint.

**Policy implication.** If Brazil's development goal is to direct capital support toward
genuinely constrained small entrepreneurs, the targeting instrument should integrate governance
structure alongside revenue-based size classification. The `owner_qualification` field is
already present in CNPJ administrative records at zero additional data collection cost.
Adding it as a screening variable would substantially improve the precision of capital-support
programs. Revenue thresholds are necessary but insufficient — and this dataset shows precisely why.
